---
title: "gloBarAnalysis"
author: "ian enochs"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

### GLOBAL BMU Analysis
This document was created to analyze NCRMP BMU data

```{r, echo=FALSE}
library(readxl)
library(dplyr)
library(magrittr)
library(ggplot2)
library(ggpubr)
library(lubridate)

# Get your favorite anesthetics predefined for all plots, so you do not have to do define it individually for each one (but call "MyTheme" when you want to apply it). This is mine:
MyTheme<-theme_bw() +  
theme(legend.position="top",
          plot.background=element_blank(),
          #axis.text.x = element_text(angle = 90, vjust = 0.5),
          panel.grid.major.y = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          legend.box.background = element_rect(),
          legend.title = element_blank(),
          panel.background =element_rect(fill = NA, 
                                         color = "black"))#

#get data
rootFolder <-getwd() #get root folder
atlanticBarData_raw <- read_excel(paste(rootFolder,"/data/NCRMP_Atlantic_allBmuData_March2024.xlsx",sep=""))#import data
pacificBarData_raw <- read_excel(paste(rootFolder,"/data/NCRMP_Pacific_allBmuData_March2024.xlsx",sep=""))#import data
allBarData_raw <- rbind(atlanticBarData_raw, pacificBarData_raw)
allBarData_clean <- allBarData_raw %>% 
  filter(fate=='SUCCESSFULLY RECOVERED') %>% #remove bars that weren't recovered
  filter(species!='Isopora sp.') %>% #remove bars made of Isopora
  filter(tagNum != 1022)%>% #remove tag 1022 that was left behind and collected the follow collection period
  select(-c(dimensions,carbMDPre, carbEQPre, carbMDPost,carbEQPre)) %>% #remove unnecessary columns
  mutate(daysDeployed = as.numeric(ymd(dateRecovery) - ymd(dateDeployment))) %>% 
  mutate(yearsDeployed=daysDeployed/365)%>% 
  mutate(basin =(case_when(
    region == "DRTO"  ~ "Atlantic", 
    region == "FGB" ~ "Atlantic",
    region == "FLK"  ~ "Atlantic",
    region == "MARIAN" ~ "Pacific",
    region == "MHI" ~ "Pacific",
    region == "PR"  ~ "Atlantic",
    region == "PRIA"  ~ "Pacific",
    region == "STJ"  ~ "Atlantic",
    region == "STT"  ~ "Atlantic",
    region == "STX"  ~ "Atlantic")))

#standardize data to surface area and days.
preSA = 24 #cm2 of exterior facing bar in a 2 x 1 x 5 cm block
allBarData_clean <- allBarData_clean  %>% 

#CHECK IF CORRECT  
  mutate(grazing = volumeBlockPost-macroboring-volumeBlockPre) %>%  #calculates grazing as volume in block change not including macroboring
  mutate(deltaMass_mgcm2y = 1000*(massCleanPost-massEpoxiedPre)/preSA/yearsDeployed) %>%  
  mutate(deltaVolume_mm3cm2y = 1000*(volumeBlockPost-volumeBlockPre)/preSA/yearsDeployed) %>%  
  mutate(deltaDensity_mgcm3cm2y = 1000*(densityCTPost-densityCTPre)/preSA/yearsDeployed) %>%  
  mutate(macroboring_mm3cm2y = 1000*macroboring/preSA/yearsDeployed) %>%  
  mutate(accretion_mm3cm2y = 1000*accretion/preSA/yearsDeployed) %>% 
  mutate(grazing_mm3cm2y = 1000*grazing/preSA/yearsDeployed)
```

## Summary of data
**Total number of BARs**
```{r, echo=FALSE}
nrow(allBarData_clean)
allBarData_clean %>%
  group_by(basin) %>%
  summarise(maxDays = max(daysDeployed), minDays = min(daysDeployed), n = n())
allBarData_clean %>%
  group_by(region) %>%
  summarise(maxDays = max(daysDeployed), minDays = min(daysDeployed), n = n())
```
<br><br>

## QA/QC of deltaMass_mgcm2y
**List tags where deltaMass_mgcm2y = "NA":**
```{r, echo=FALSE}
  allBarData_clean$tagNum[which(is.na(allBarData_clean$deltaMass_mgcm2y))]
```
<br><br>
**Historgram of deltaMass_mgcm2y values**
```{r, echo=FALSE}
  ggplot(allBarData_clean, aes(x = deltaMass_mgcm2y)) +
  geom_histogram(binwidth = 5, color = "black", fill = "white") +
  labs(title = "Histogram of deltaMass_mgcm2y", x = "deltaMass_mgcm2y", y = "Frequency")
```

## QA/QC of deltaVolume_mm3cm2y
**List tags where deltaVolume_mm3cm2y = "NA":**
```{r, echo=FALSE}
allBarData_clean$tagNum[which(is.na(allBarData_clean$deltaVolume_mm3cm2y))]
```
<br><br>
**Historgram of deltaVolume_mm3cm2y values**
```{r, echo=FALSE}
  ggplot(allBarData_clean, aes(x = deltaVolume_mm3cm2y)) +
  geom_histogram(binwidth = 5, color = "black", fill = "white") +
  labs(title = "Histogram of deltaVolume_mm3cm2y", x = "deltaVolume_mm3cm2y", y = "Frequency")
```

## QA/QC of deltaDensity_mgcm3cm2y
**List tags where deltaDensity_mgcm3cm2y = "NA":**
```{r checkNADensity, echo=FALSE}
allBarData_clean$tagNum[which(is.na(allBarData_clean$deltaDensity_mgcm3cm2y))]
```
<br><br>
**Historgram of deltaDensity_mgcm3cm2y values**
```{r, echo=FALSE}
  ggplot(allBarData_clean, aes(x = deltaDensity_mgcm3cm2y)) +
  geom_histogram(binwidth = .1, color = "black", fill = "white") +
  labs(title = "Histogram of deltaDensity_mgcm3cm2y", x = "deltaDensity_mgcm3cm2y", y = "Frequency")
```


## QA/QC of macroboring_mm3cm2y
**List tags where macroboring_mm3cm2y = "NA":**
```{r checkNAMacroboring, echo=FALSE}
allBarData_clean$tagNum[which(is.na(allBarData_clean$macroboring_mm3cm2y))]
```
<br><br>
**Historgram of macroboring_mm3cm2y values**
```{r, echo=FALSE}
  ggplot(allBarData_clean, aes(x = macroboring_mm3cm2y)) +
  geom_histogram(binwidth = 1, color = "black", fill = "white") +
  labs(title = "Histogram of macroboring_mm3cm2y", x = "macroboring_mm3cm2y", y = "Frequency")
```

## QA/QC of accretion_mm3cm2y
**List tags where accretion_mm3cm2y = "NA":**
```{r checkNAAccretion, echo=FALSE}
allBarData_clean$tagNum[which(is.na(allBarData_clean$accretion_mm3cm2y))]
```
<br><br>
**Historgram of accretion_mm3cm2y values**
```{r, echo=FALSE}
  ggplot(allBarData_clean, aes(x = accretion_mm3cm2y)) +
  geom_histogram(binwidth = 5, color = "black", fill = "white") +
  labs(title = "Histogram of accretion_mm3cm2y", x = "accretion_mm3cm2y", y = "Frequency")
```

## QA/QC of grazing_mm3cm2y
<br><br>
**List tags where grazing_mm3cm2y = "NA":**
```{r, echo=FALSE}
allBarData_clean$tagNum[which(is.na(allBarData_clean$grazing_mm3cm2y))]
```
<br><br>
**List tags where grazing_mm3cm2y > 0 and list of those values:**
```{r, echo=FALSE}
allBarData_clean$tagNum[which(allBarData_clean$grazing_mm3cm2y>0)]
allBarData_clean$grazing_mm3cm2y[which(allBarData_clean$grazing_mm3cm2y>0)]
```
<br><br>
**Historgram of grazing_mm3cm2y values**
```{r, echo=FALSE}
  ggplot(allBarData_clean, aes(x = grazing_mm3cm2y)) +
  geom_histogram(binwidth = 5, color = "black", fill = "white") +
  labs(title = "Histogram of grazing_mm3cm2y", x = "grazing_mm3cm2y", y = "Frequency")
```

## General trends
**Plot of deltaVolume_mm3cm2y vs. deltaMass_mgcm2y:**
```{r, echo=FALSE}

p <- ggplot(allBarData_clean, aes(x = deltaMass_mgcm2y, y = deltaVolume_mm3cm2y)) +
  geom_point(aes(color = factor(region))) +  # Scatterplot
  geom_smooth(method = "lm", se = TRUE) +  # Regression line
  theme_minimal()

# Calculate R-squared value
lm_model_volVsMass <- lm(deltaVolume_mm3cm2y ~ deltaMass_mgcm2y, data = allBarData_clean)
r_squared_volVsMass <- summary(lm_model_volVsMass)$r.squared


p + geom_text(data = subset(allBarData_clean, deltaMass_mgcm2y > 100), aes(label = tagNum), 
            hjust = -0.1, vjust = -0.5, size = 2)+# Add tag numbers here delta mass is greater than 100
    geom_text(x = 200, y = -100, label = paste("R^2 =", round(r_squared_volVsMass, 2)), size = 3)# Add R-squared value to the plot
```
<br><br>
**Plot of deltaMass_mgcm2y vs. daysDeployed:**
```{r, echo=FALSE}

plotDeltaMassVsDays <- ggplot(allBarData_clean, aes(x = daysDeployed, y = deltaMass_mgcm2y)) +
  geom_point(aes(color = factor(region))) +  # Scatterplot
  geom_smooth(method = "lm", se = TRUE) +  # Regression line
  theme_minimal()

# Calculate R-squared value
lm_model_massVsDays <- lm(deltaMass_mgcm2y ~ daysDeployed, data = allBarData_clean)
r_squared_massVsDays <- summary(lm_model_massVsDays)$r.squared

# Add R-squared value to the plot
plotDeltaMassVsDays + geom_text(data = subset(allBarData_clean, daysDeployed > 2500), aes(label = tagNum), hjust = -0.1, vjust = -0.5, size = 2)+# Add tag numbers here where days deployed is greater than 2500
    geom_text(data = subset(allBarData_clean, deltaMass_mgcm2y > 100), aes(label = tagNum), hjust = -0.1, vjust = -0.5, size = 2)+# Add tag numbers here where delta mass is greater than 100
      geom_text(x = 2000, y = 100, label = paste("R^2 =", round(r_squared_massVsDays, 2)), size = 3)
```
<br><br>
**Plot of macroboring_mm3cm2y vs. daysDeployed:**
```{r, echo=FALSE}
# Calculate R-squared value
lm_model_macroboringVsDays <- lm(macroboring_mm3cm2y ~ daysDeployed, data = allBarData_clean)
r_squared_macroboringVsDays <- summary(lm_model_macroboringVsDays)$r.squared

plotMacroboringVsDays <- ggplot(allBarData_clean, aes(x = daysDeployed, y = macroboring_mm3cm2y)) +
  geom_point(aes(color = factor(region))) +  # Scatterplot
  geom_smooth(method = "lm", se = TRUE) +  # Regression line
  theme_minimal()+
  geom_text(x = 2000, y = -30, label = paste("R^2 =", round(r_squared_macroboringVsDays, 2)), size = 3)
plotMacroboringVsDays
```
<br><br>
**Plot of grazing_mm3cm2y vs. daysDeployed:**
```{r, echo=FALSE}
# Calculate R-squared value
lm_model_grazingVsDays <- lm(grazing_mm3cm2y ~ daysDeployed, data = allBarData_clean)
r_squared_grazingVsDays <- summary(lm_model_grazingVsDays)$r.squared

plotGrazingVsDays <- ggplot(allBarData_clean, aes(x = daysDeployed, y = grazing_mm3cm2y)) +
  geom_point(aes(color = factor(region))) +  # Scatterplot
  geom_smooth(method = "lm", se = TRUE) +  # Regression line
  theme_minimal()+
  geom_text(x = 2000, y = -75, label = paste("R^2 =", round(r_squared_grazingVsDays, 2)), size = 3)
plotGrazingVsDays
```
<br><br>
**Plot of accretion_mm3cm2y vs. daysDeployed:**
```{r, echo=FALSE}
# Calculate R-squared value
lm_model_accretionVsDays <- lm(accretion_mm3cm2y ~ daysDeployed, data = allBarData_clean)
r_squared_accretionVsDays <- summary(lm_model_accretionVsDays)$r.squared

plotAccretionVsDays <- ggplot(allBarData_clean, aes(x = daysDeployed, y = accretion_mm3cm2y)) +
  geom_point(aes(color = factor(region))) +  # Scatterplot
  geom_smooth(method = "lm", se = TRUE) +  # Regression line
  theme_minimal()+
  geom_text(x = 2000, y = 100, label = paste("R^2 =", round(r_squared_accretionVsDays, 2)), size = 3)
plotAccretionVsDays
```
<br><br>


**Grazing vs. Accretion <br>**
```{r, echo=FALSE}

grazingVsAccretion <- ggplot(allBarData_clean, aes(x = accretion_mm3cm2y, y = grazing_mm3cm2y)) +
  geom_point(aes(color = factor(region))) +  # Scatterplot
  geom_smooth(method = "lm", se = TRUE) +  # Regression line
  theme_minimal()
grazingVsAccretion
```

<br><br>
**deltaMass_mgcm2y vs. grazing_mm3cm2y <br>**
```{r, echo=FALSE}

deltaMassVsGrazing <- ggplot(allBarData_clean, aes(x = grazing_mm3cm2y, y = deltaMass_mgcm2y)) +
  geom_point(aes(color = factor(region))) +  # Scatterplot
  geom_smooth(method = "lm", se = TRUE) +  # Regression line
  theme_minimal()
deltaMassVsGrazing

deltaMassGrazingStats <- allBarData_clean %>%
  filter(grazing_mm3cm2y != "")%>% #remove blank grazing values
  filter(region != "STT")%>% #remove STT where only one
  group_by(basin, region, siteName) %>%
  summarise(meanGrazing = mean(grazing_mm3cm2y), sdGrazing = sd(grazing_mm3cm2y), seGrazing = (sd(grazing_mm3cm2y)/sqrt(n())), meanDeltaMass = mean(deltaMass_mgcm2y), sdDeltaMass = sd(deltaMass_mgcm2y), seDeltaMass = (sd(deltaMass_mgcm2y)/sqrt(n())), n = n())



deltaMassGrazingStats
deltaMassGrazingStatsPlot <- ggplot(deltaMassGrazingStats, aes(x = meanGrazing, y = meanDeltaMass)) +
  geom_point(aes(shape = basin, size = 1, color = factor(region))) +  # Scatterplot
  geom_errorbar(aes(ymin=meanDeltaMass-seDeltaMass, ymax=meanDeltaMass+seDeltaMass), width=.1) +
  geom_errorbar(aes(xmin=meanGrazing-seGrazing, xmax=meanGrazing+seGrazing), width=.1) +
  geom_smooth(method = "lm", se = TRUE) +  # Regression line
  theme_minimal()
deltaMassGrazingStatsPlot



```





## Atlantic and Caribbean: 




**deltaMass_mgcm2y boxplots by region and subregion**
```{r, echo=FALSE}
allBarData_clean$subRegion <- factor(allBarData_clean$subRegion, levels = c("north", "east", "south", "west", "upper","middle","lower"))
allBarData_clean$subRegion2 <- paste(allBarData_clean$region, allBarData_clean$subRegion,sep="-")

plot_deltaMassRegion_Atlantic <- ggplot(data=allBarData_clean[!is.na(allBarData_clean$subRegion),],
                                    aes(x=region, fill=subRegion2,
                                        y=deltaMass_mgcm2y)) +
  geom_boxplot(#width = 0.5, color = "black",
               position = position_dodge2(preserve = "single")) +
  geom_hline(yintercept=0, linetype="dashed", color = "red")+
  #facet_wrap(~region, scales = "free_x") + MyTheme +
  scale_y_continuous(name=(expression(~Delta~mass~(mg~cm^-2~y^-1))))+
  scale_x_discrete(name="")+MyTheme
plot_deltaMassRegion_Atlantic
```
<br><br>
**Macroboring boxplots by region and subregion**
```{r, echo=FALSE}
plot_macroboringRegion_Atlantic <- ggplot(data=allBarData_clean[!is.na(allBarData_clean$subRegion),],
                                    aes(x=region, fill=subRegion2,
                                        y=macroboring_mm3cm2y)) +
  geom_boxplot(#width = 0.5, color = "black",
               position = position_dodge2(preserve = "single")) +
  geom_hline(yintercept=0, linetype="dashed", color = "red")+
  #facet_wrap(~region, scales = "free_x") + MyTheme +
  scale_y_continuous(name=(expression(~Delta~volume~(mm^3~cm^-2~y^-1))))+
  scale_x_discrete(name="")+MyTheme
plot_macroboringRegion_Atlantic
```
<br><br>
**Grazing boxplots by region and subregion<br>**
*Note that the initial volumes are missing for some USVI and PR data, thereby removing some of those grazing values.
```{r, echo=FALSE}
plot_grazingRegion_Atlantic <- ggplot(data=allBarData_clean[!is.na(allBarData_clean$subRegion),],
                                    aes(x=region, fill=subRegion2,
                                        y=grazing_mm3cm2y)) +
  geom_boxplot(#width = 0.5, color = "black",
               position = position_dodge2(preserve = "single")) +
  geom_hline(yintercept=0, linetype="dashed", color = "red")+
  #facet_wrap(~region, scales = "free_x") + MyTheme +
  scale_y_continuous(name=(expression(~Delta~volume~(mm^3~cm^-2~y^-1))))+
  scale_x_discrete(name="")+MyTheme
plot_grazingRegion_Atlantic
```
<br><br>
**Accretion boxplots by region and subregion<br>**
```{r, echo=FALSE}
plot_accretionRegion_Atlantic <- ggplot(data=allBarData_clean[!is.na(allBarData_clean$subRegion),],
                                    aes(x=region, fill=subRegion2,
                                        y=accretion_mm3cm2y)) +
  geom_boxplot(#width = 0.5, color = "black",
               position = position_dodge2(preserve = "single")) +
  geom_hline(yintercept=0, linetype="dashed", color = "red")+
  #facet_wrap(~region, scales = "free_x") + MyTheme +
  scale_y_continuous(name=(expression(~Delta~volume~(mm^3~cm^-2~y^-1))))+
  scale_x_discrete(name="")+MyTheme
plot_accretionRegion_Atlantic

```

