---
title: "globarAnalysisV2"
author: "ian enochs"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

This document was created to analyze NCRMP BMU data

# Load and clean data
### Load Packages
```{r loadPackages, echo=FALSE}
# Core data stack
library(tidyverse)   # dplyr, ggplot2, tidyr, readr, purrr, tibble, stringr, forcats, magrittr
library(lubridate)
library(readxl)
library(jsonlite)
library(curl)
library(readr)

# Spatial
library(sf)          # vector
library(terra)       # raster/terra
library(units)
library(rnaturalearth)
library(rnaturalearthdata)
library(leaflet)
library(mapview)

# Viz helpers & layout
library(ggrepel)
library(patchwork)
library(cowplot)
library(ggpubr)

# Stats / ecology
library(vegan)
library(cluster)
library(lme4)
library(lmerTest)
library(emmeans)
library(dplyr)
library(tidyr)
library(broom)

# Reporting
library(knitr)
library(kableExtra)
library(rmarkdown)
```

### Set aesthetics
```{r setAesthetics, echo=FALSE}
# Create color scale for loc1.2 categories
loc_colors <- c(
  # Caribbean
    "caribbean.puertoRico"   = "#FFD34D",
    "caribbean.saintCroix"   = "#FF9B00",
    "caribbean.saintJohn"    = "#FF6B00",
    "caribbean.saintThomas"  = "#CC4E00",
  # Florida
    "florida.keys"           = "#660000",
    "florida.dryTortugas"    = "#E04B3B",
  # Flower Garden Banks 
    "flowerGardenBanks.flowerGardenBanks" = "#7B4F1D",
  # Marianas
    "marianas.guam"   = "#C51B7D",
    "marianas.maug"   = "#882E72",
    "marianas.pagan"  = "#5E4FA2",
    "marianas.saipan" = "#2A1E5C",
  # Middle Hawaiian Islands
    "middleHawaiianIslands.hawaii" = "#005B5A",
    "middleHawaiianIslands.oahu"   = "#4ED1C5",
  # Northwest Hawaiian Islands
    "northwestHawaiianIslands.frenchFrigateShoals" = "#004B9A",
    "northwestHawaiianIslands.pearlHermes"         = "#6FB6FF",
  # Pacific Remote Island Area
    "pacificRemoteIslandArea.jarvis"  = "#003B1A",
    "pacificRemoteIslandArea.palmyra" = "#1E8E3E",
    "pacificRemoteIslandArea.wake"    = "#7BC96F",
  # Samoa
    "samoa.ofu"     = "#1C1C1C",
    "samoa.tau"     = "#7A7A7A",
    "samoa.tutuila" = "#C8C8C8"
)
```

### Define data import paths
```{r dataImportPaths, echo=FALSE}
#Set root folder and working directories
rootFolder <- file.path("C:/Users/ian.enochs/Documents/GitHub/globalBar/analysis")  # Attempt Ian's working directory first
if (!dir.exists(rootFolder)) {                                                      # If Ian's path does not exist, use John's path instead
  rootFolder <- file.path("/Users/john.morris/Documents/GitHub/globalBar/analysis")
}
setwd(rootFolder)

#Bar directory and paths
data_dir   <- file.path(rootFolder, "data")
atl_path   <- file.path(data_dir, "NCRMP_Atlantic_allBmuData.xlsx")
pac_path   <- file.path(data_dir, "NCRMP_Pacific_allBmuData.xlsx")
dec2025Update_path   <- file.path(data_dir, "Pac_Atl_NCRMP_2024_updateDec2025.xlsx") #dec2025 update path

#Parrotfish directory and paths
p_dir <- file.path(rootFolder, "data", "parrotfish")
parrot_path <- file.path(p_dir, "BMU_Parrotfish_AnalysisReady.csv")

#Environmental directory and path
e_dir <- file.path(rootFolder, "data", "environmental")
env_path <- file.path(e_dir, "eds_time.csv")
```

### Create functions for importing data
```{r dataImportFunctions, echo=FALSE}
#Function to remove rows if all cells are NA
drop_all_na_rows <- function(df) {
  df %>%
    dplyr::mutate(dplyr::across(dplyr::where(is.character), ~ dplyr::na_if(trimws(.), ""))) %>%
    dplyr::filter(!dplyr::if_all(dplyr::everything(), ~ is.na(.)))
}

#Function to assign basins based on regions
to_basin <- function(region) dplyr::case_when(
  region %in% c("DRTO","FGB","FLK","PR","STJ","STT","STX") ~ "atlantic",
  region %in% c("MARIAN","MHI","NWHI","PRIA","SAMOA")      ~ "pacific",
  TRUE ~ NA_character_
)

#Function to assign loc1 based on region
to_loc1 <- function(region) dplyr::case_when(
  region %in% c("DRTO","FLK")           ~ "florida",
  region == "FGB"                       ~ "flowerGardenBanks",
  region %in% c("PR","STJ","STT","STX") ~ "caribbean",
  region == "MARIAN"                    ~ "marianas",
  region == "MHI"                       ~ "middleHawaiianIslands",
  region == "NWHI"                      ~ "northwestHawaiianIslands",
  region == "PRIA"                      ~ "pacificRemoteIslandArea",
  region == "SAMOA"                     ~ "samoa",
  TRUE ~ NA_character_
)

#Function to assign loc2 based on region and subregion
to_loc2 <- function(region, subRegion) dplyr::case_when(
  region == "DRTO" ~ "dryTortugas",
  region == "FGB"  ~ "flowerGardenBanks",
  region == "FLK"  ~ "keys",
  region == "PR"   ~ "puertoRico",
  region == "STJ"  ~ "saintJohn",
  region == "STT"  ~ "saintThomas",
  region == "STX"  ~ "saintCroix",
  subRegion == "OFU" ~ "ofu",
  subRegion == "PAL" ~ "palmyra",
  subRegion == "TAU" ~ "tau",
  subRegion == "TUT" ~ "tutuila",
  subRegion == "WAK" ~ "wake",
  subRegion == "JAR" ~ "jarvis",
  subRegion == "HAW" ~ "hawaii",
  subRegion == "OAH" ~ "oahu",
  subRegion == "FFS" ~ "frenchFrigateShoals",
  subRegion == "PHR" ~ "pearlHermes",
  subRegion == "GUA" ~ "guam",
  subRegion == "SAI" ~ "saipan",
  subRegion == "PAG" ~ "pagan",
  subRegion == "MAU" ~ "maug",
  TRUE ~ NA_character_
)

#FUNCTIONS TO ASSIGN LOC3
#Shared helpers
safe_concat <- function(l2, subR) {
  s <- ifelse(is.na(subR) | trimws(subR) == "", NA, subR)
  ifelse(is.na(s), l2, paste(l2, s, sep = "-"))
}
first3_letters_lower <- function(x) {
  tolower(substr(gsub("\\s+", "", x), 1, 3))
}
normalize_site <- function(siteName) {
  ifelse(is.na(siteName), "", gsub("_", "-", siteName, fixed = TRUE))
}

#Pacific site name map
  pac_dir_map <- c(
    # Tutuila
    "OCC-TUT-002"="north","OCC-TUT-007"="east","OCC-TUT-017"="south","OCC-TUT-022"="west",
    # Ofu
    "OCC-OFU-002"="north","OCC-OFU-009"="south",
    # Tau
    "OCC-TAU-002"="north","OCC-TAU-011"="west",
    # Jarvis
    "OCC-JAR-003"="east","OCC-JAR-008"="south","OCC-JAR-013"="west",
    # Palmyra
    "OCC-PAL-003"="north","OCC-PAL-008"="east","OCC-PAL-017"="west",
    # Hawaii
    "OCC-HAW-014"="north","OCC-HAW-002"="north","OCC-HAW-006"="east","OCC-HAW-066"="east","OCC-HAW-008"="south","OCC-HAW-010"="west",#note east is listed as 066 in parrotfish and 006 in bar
    # Oahu
    "OCC-OAH-002"="north","OCC-OAH-005"="east","OCC-OAH-007"="east","OCC-OAH-008"="southeast", "OCC-OAH-009"="south","OCC-OAH-010"="south","OCC-OAH-014"="southwest","OCC-OAH-012"="west",#note 009 doesn't exist
    # FFS
    "OCC-FFS-002"="north","OCC-FFS-004"="north","OCC-FFS-007"="east","OCC-FFS-010"="south","OCC-FFS-014"="west",
    # Pearl & Hermes
    "OCC-PHR-002"="north","OCC-PHR-016"="west",                  "OCC-PHR-007"="east","OCC-PHR-012"="south",
    # Wake
    "OCC-WAK-002"="north","OCC-WAK-005"="east","OCC-WAK-008"="south","OCC-WAK-013"="west",
    # Guam
    "OCC-GUA-002"="north","OCC-GUA-005"="east","OCC-GUA-009"="south","OCC-GUA-015"="west",
    # Saipan
    "OCC-SAI-002"="north","OCC-SAI-005"="east","OCC-SAI-009"="south","OCC-SAI-012"="west","OCC-SAI-016"="northwest",
    # Pagan
    "OCC-PAG-002"="north","OCC-PAG-006"="east","OCC-PAG-009"="south","OCC-PAG-013"="west",
    # Maug
    "OCC-MAU-002"="north","OCC-MAU-005"="east","OCC-MAU-019"="south","OCC-MAU-008"="west","OCC-MAU-016"="innerHCO2","OCC-MAU-018"="innerMCO2"
  )

to_loc3_bar <- function(region, loc2, subRegion, siteName, basin) {
  bsn       <- tolower(basin)
  site_norm <- normalize_site(siteName)
  dplyr::case_when(
    # Puerto Rico
    loc2 == "puertoRico" & siteName == "Fajardo" ~ "puertoRico-northeast",
    bsn == "atlantic" & loc2 == "puertoRico" & siteName %in% c("laParguera","La Parguera") ~
      "puertoRico-southwest",

    # Florida Keys (BAR logic)
    loc2 == "keys" & siteName == "UKI1" ~ "keys-uki1",
    loc2 == "keys" & siteName == "UKI2" ~ "keys-uki2",
    loc2 == "keys" & siteName == "Cheeca Rocks" ~ "keys-uki3",
    loc2 == "keys" & siteName == "UKO1" ~ "keys-uko", #changed to uko to combine uko1 and 3    
    loc2 == "keys" & siteName == "UKO3" ~ "keys-uko", #changed to uko to combine uko1 and 3   
    loc2 == "keys" & siteName == "MKI1" ~ "keys-mki1",    
    loc2 == "keys" & siteName == "MKI2" ~ "keys-mki2",
    loc2 == "keys" & siteName == "MKO1" ~ "keys-mko1",
    loc2 == "keys" & siteName == "MKO2" ~ "keys-mko2",
    loc2 == "keys" & siteName == "lKI1" ~ "keys-lki1",
    loc2 == "keys" & siteName == "lKI2" ~ "keys-lki2",    
    loc2 == "keys" & siteName == "lKo1" ~ "keys-lko1",
    loc2 == "keys" & siteName == "lKo2" ~ "keys-lko2",        
    
    # Pacific mapped sites
    bsn == "pacific" & site_norm %in% names(pac_dir_map) ~
      paste(loc2, pac_dir_map[site_norm], sep = "-"),

    bsn == "pacific" ~ loc2,

    # Atlantic general
    bsn == "atlantic" & loc2 %in% c(
      "dryTortugas","flowerGardenBanks","puertoRico",
      "saintJohn","saintThomas","saintCroix"
    ) ~ safe_concat(loc2, subRegion),

    TRUE ~ NA_character_
  )
}  

to_loc3_parrot <- function(region, loc2, subRegion, siteName, basin) {
  site_norm <- normalize_site(siteName)

  dplyr::case_when(
    # Puerto Rico overrides
    loc2 == "puertoRico" & siteName == "Fajardo"  ~ "puertoRico-northeast",
#    loc2 == "puertoRico" & siteName == "fajardo" & subRegion == "east"  ~ "puertoRico-northeast",
#    loc2 == "puertoRico" & siteName == "fajardo" & subRegion == "north" ~ "puertoRico-north",
    loc2 == "puertoRico" & siteName == "culebra" & subRegion == "east"  ~ "puertoRico-east",
    loc2 == "puertoRico" & siteName == "laParguera" & subRegion == "south" ~ "puertoRico-southwest",
    
    # Florida Keys (PARROT logic)
    basin == "atlantic" & loc2 == "keys" & subRegion == "lower"  & site_norm == "inshore"  ~ "keys-lki",
    basin == "atlantic" & loc2 == "keys" & subRegion == "lower"  & site_norm == "offshore" ~ "keys-lko",
    basin == "atlantic" & loc2 == "keys" & subRegion == "middle" & site_norm == "inshore"  ~ "keys-mki",
    basin == "atlantic" & loc2 == "keys" & subRegion == "middle" & site_norm == "offshore" ~ "keys-mko",
    basin == "atlantic" & loc2 == "keys" & subRegion == "upper"  & site_norm == "inshore"  ~ "keys-uki",
    basin == "atlantic" & loc2 == "keys" & subRegion == "upper"  & site_norm == "offshore" ~ "keys-uko",

    # Pacific mapped sites
    basin == "pacific" & site_norm %in% names(pac_dir_map) ~
      paste(loc2, pac_dir_map[site_norm], sep = "-"),

    basin == "pacific" ~ loc2,

    # Atlantic general
    basin == "atlantic" & loc2 %in% c(
      "dryTortugas","flowerGardenBanks","puertoRico",
      "saintJohn","saintThomas","saintCroix"
    ) ~ safe_concat(loc2, subRegion),

    TRUE ~ NA_character_
  )
}  

to_loc3_env <- function(loc2, lat) {
  env_lat_map <- c(
    # puerto rico
    "18.48227"="north","18.33066"="east","18.33696"="northeast","17.91047"="south","17.95355"="southwest","18.34435"="west",
    # saint croix
    "17.82805"="east","17.78532"="north","17.70121"="south","17.7338"="west",
    # saint john
    "18.36316"="north","18.34744"="east","18.32802"="west","18.30976"="south",
    # saint thomas
    "18.34402"="south","18.38262"="north","18.34082"="west",
    # dry tortugas
    "24.61123"="west",
    # florida keys NOTE: Only using the uko3 for the analysis because I've used the 24.88266 uko3 site to change to uko, which will align with other datasets loc3 levels
    "24.5859"="lki","24.57899"="lki","24.52202"="lko","24.55326"="lko","24.68554"="mki","24.68988"="mki","24.63127"="mko","24.63388"="mko","24.95375"="uki","24.9025"="uko1","24.88266"="uko","24.89742"="uki",
    # flower garden banks
    "27.9074"="east",
    # guam
    "13.6033"="north","13.47516"="east","13.24126"="south","13.2421"="OCC-GUA-009","13.52894"="west",
    # maug
    "20.0357"="north","20.02033"="east","20.01312"="west","20.02296"="innerHCO2","20.02249"="innerMCO2","20.01398"="south",
    # pagan
    "18.16784"="north","18.09612"="east","18.04564"="south","18.12014"="west",
    # saipan
    "15.29062"="north","15.25866"="east","15.09785"="south","15.15639"="OCC-SAI-012","15.15645"="west","15.25719"="northwest",
    # hawaii
    "20.26828"="north","20.24913"="northExtra","19.48642"="east", "19.06075"="south", "19.24432"="west", #removing double north site for consistency
    # oahu
    "21.45862"="east","21.47972"="eastExtra","21.28838"="south","21.26184"="southeast","21.53412"="west","21.30566"="southwest", #removing double east site for consistency
    # french frigate shoals
    "23.87806"="north","23.88021"="north","23.83515"="east","23.63686"="south","23.79247"="west",
    # pearl and hermes
    "27.94055"="north","27.95493"="north","27.86679"="east","27.78237"="south","27.79316"="OCC-PHR-016",
    # jarvis
    "-0.36896"="west","-0.38234"="south",
    # palmyra
    "5.89733"="north","5.86379"="east","5.88322"="west",
    # wake
    "19.31602"="north","19.30616"="east","19.28065"="south","19.31625"="west",
    # ofu
    "-14.15785"="north","-14.1733"="south",
    # tau
    "-14.2124"="north","-14.25755"="west",
    # tutuila
    "-14.24479"="north","-14.28521"="east","-14.36351"="south","-14.29464"="west"
  )
  # Make sure map keys are consistently 5-decimal strings
  nm <- suppressWarnings(as.numeric(names(env_lat_map)))
  names(env_lat_map) <- sprintf("%.5f", nm)

  # Coerce lat safely + format the same way
  lat_num <- suppressWarnings(as.numeric(lat))
  key <- sprintf("%.5f", round(lat_num, 5))

  suffix <- unname(env_lat_map[key])

  # Build loc3 = loc2-suffix (only when suffix exists)
  out <- ifelse(is.na(suffix) | is.na(loc2) | trimws(loc2) == "",
                NA_character_,
                paste0(loc2, "-", suffix))

  out
}
```  

### Read and bind raw data
```{r readBindRawData, echo=FALSE}
#FUNCTION FOR READING DATA TYPES
# For Excel (.xlsx)
read_bind_excel <- function(paths, sheet = NULL) {
  purrr::map_dfr(
    paths,
    ~ readxl::read_excel(.x, sheet = sheet)
  )
}
# For CSVs
read_bind_csv <- function(paths, col_types) {
  purrr::map_dfr(paths, ~ readr::read_csv(.x, col_types = col_types, show_col_types = FALSE))
}
ct <- cols(
  .default       = col_guess(),
  yearFishSurvey = col_integer(),
  siteFishSurvey = col_character()
)

#READ AND BIND DATA
# Read & Bind Bar Data
barData_raw <-
  read_bind_excel(c(atl_path, pac_path, dec2025Update_path)) %>%
  drop_all_na_rows()

# Read & Bind Parrot Community Data
parrotData_raw <- read_bind_csv(parrot_path, ct) %>%
  drop_all_na_rows()

# Read & bind environmental data
environmentalData_raw <- read_bind_csv(c(env_path), cols(.default = col_guess())) %>%
  drop_all_na_rows()
```

### Fix inconsistencies in the raw data
```{r fixRawDataConsistencies, echo=FALSE}
#barData QAQC
#Fix typos in barData site names
barData <- barData_raw %>%
  mutate(
    siteName = case_when(
      siteName == "OCC-HAW-10" ~ "OCC-HAW-010",
      siteName == "OCC-OAH-012" & as.character(tagNum) == "1660" ~ "OCC-OAH-002",
      siteName == "OCC-JAR-003" & tagNum %in% c(3265, 3266, 3267, 3268) ~ "OCC-JAR-013",
      TRUE ~ siteName
          )
  )
#Species naming consistencies 
barData <- barData %>%
  mutate(
    species = case_when(
      species %in% c("O. faveolata", "Orbicella faveolata", "Orbicella sp.", "O.faveolata") ~ "ofav",
      species %in% c("P. lobata", "Porites lobata") ~ "plob",
      species %in% c("Isopora sp.") ~ "iso",
      TRUE ~ species
    )
  )
other_species <- barData %>%
  distinct(species) %>%
  arrange(species)
#List of all species
other_species

#Parrot Data QAQC
parrotData <- parrotData_raw

#Environmental Data QAQC
environmentalData <- environmentalData_raw
```

### Create BAR Loc fields
```{r creatLocFields, echo=FALSE}
# Create BAR loc fields
barData <- barData %>%
  select(-c(dimensions, carbMDPre, carbEQPre, carbMDPost)) %>%
  mutate(
    daysDeployed  = as.numeric(ymd(dateRecovery) - ymd(dateDeployment)),
    yearsDeployed = daysDeployed / 365
  ) %>%
  mutate(
    basin = to_basin(region),
    loc1  = to_loc1(region),
    loc2  = to_loc2(region, subRegion),
    loc3  = to_loc3_bar(region, loc2, subRegion, siteName, basin),
    loc1.2 = paste(loc1, loc2, sep = ".")
  )

# Create parrot community loc fields
parrotData <- parrotData %>%
  mutate(
    basin = to_basin(region),
    loc1  = to_loc1(region),
    loc2  = to_loc2(region, subRegion),
    loc3  = to_loc3_parrot(region, loc2, subRegion, siteName, basin),
    loc1.2 = paste(loc1, loc2, sep = ".")
  )

# Create environmental data loc fields
environmentalData <- environmentalData %>%
  mutate(
    loc3   = to_loc3_env(loc2, lat),
    loc1.2 = paste(loc1, loc2, sep = ".")
  )
```

### Fix depth issues
```{r fixDepths, echo=FALSE}
#Fix depths based on csv tables
#Pacific depths (match by siteName)
pac_depths <- read_csv(file.path(data_dir, "barData_pac_depths.csv"), show_col_types = FALSE) %>%
  transmute(
    siteName = str_trim(siteName),
    depth_new = depth    # <-- change if your CSV column isn't literally named "depth"
  )
barData <- barData %>%
  mutate(
    siteName = str_trim(siteName),
    basin_l  = tolower(basin)
  ) %>%
  left_join(pac_depths, by = "siteName") %>%
  mutate(
    depth = if_else(basin_l == "pacific" & !is.na(depth_new), depth_new, depth)
  ) %>%
  select(-depth_new, -basin_l)

# Atlantic depths (match by loc3)
atl_depths <- read_csv(file.path(data_dir, "barData_atl_depths.csv"), show_col_types = FALSE) %>%
  transmute(
    loc3 = str_trim(loc3),
    depth_new = depth    # <-- change if your CSV column isn't literally named "depth"
  )
barData <- barData %>%
  mutate(
    loc3    = str_trim(loc3),
    basin_l = tolower(basin)
  ) %>%
  left_join(atl_depths, by = "loc3") %>%
  mutate(
    depth = if_else(basin_l == "atlantic" & !is.na(depth_new), depth_new, depth)
  ) %>%
  select(-depth_new, -basin_l)

#Check completeness by creating list of unmatched Pacific siteName
pac_sites_in_bar <- barData %>%
  filter(tolower(basin) == "pacific") %>%
  distinct(siteName) %>%
  filter(!is.na(siteName), siteName != "")
pac_sites_in_csv <- pac_depths %>%
  distinct(siteName) %>%
  filter(!is.na(siteName), siteName != "")
unmatchedDepth_pacific <- list(
  in_barData_pacific_but_not_in_csv = sort(setdiff(pac_sites_in_bar$siteName, pac_sites_in_csv$siteName)),
  in_csv_but_not_in_barData_pacific = sort(setdiff(pac_sites_in_csv$siteName, pac_sites_in_bar$siteName))
)
unmatchedDepth_pacific
#Check completeness by creating list of unmatched Atlantic loc3
atl_loc3_in_bar <- barData %>%
  filter(tolower(basin) == "atlantic") %>%
  distinct(loc3) %>%
  filter(!is.na(loc3), loc3 != "")

atl_loc3_in_csv <- atl_depths %>%
  distinct(loc3) %>%
  filter(!is.na(loc3), loc3 != "")
unmatchedDepth_atlantic <- list(
  in_barData_atlantic_but_not_in_csv = sort(setdiff(atl_loc3_in_bar$loc3, atl_loc3_in_csv$loc3)),
  in_csv_but_not_in_barData_atlantic = sort(setdiff(atl_loc3_in_csv$loc3, atl_loc3_in_bar$loc3))
)
unmatchedDepth_atlantic
```

### Add in skeletal source info
```{r add skeletal source to BMU data, echo=FALSE}
# Load BMU database (lookup table)
bmuDatabase <- read_csv(file.path(data_dir, "NCRMP_BMU_DatabaseNewFeb2025.csv"), show_col_types = FALSE)

# Join FromColony onto barData as skeletalSource (Tag_num -> tagNum)
barData <- barData %>%
  dplyr::mutate(tagNum_chr = as.character(tagNum)) %>%   # TEMP join key
  dplyr::left_join(
    bmuDatabase %>%
      dplyr::transmute(
        tagNum_chr      = as.character(Tag_num),
        skeletalSource  = as.character(FromColony)
      ),
    by = "tagNum_chr"
  ) %>%
  dplyr::select(-tagNum_chr) 

cat("Non-missing skeletalSource:",
    sum(!is.na(barData$skeletalSource)), "\n")

# See which tags failed to match
barData %>%
  filter(is.na(skeletalSource)) %>%
  distinct(tagNum) %>%
  arrange(tagNum) %>%
  print(n = 20)
```

### Calculate BAR rates
```{r calculateBarRates, echo=FALSE}
# Derived metrics (area/time standardized)
preSA <- 24  # cm^2 exterior-facing bar (2 x 1 x 5 cm block)
barData <- barData %>%
  mutate(
    grazing                = volumeBlockPost - macroboring - volumeBlockPre,
    deltaMass_mgcm2y       = 1000 * (massCleanPost - massEpoxiedPre) / preSA / yearsDeployed,
    deltaMass_kgm2y        = 0.01 * 1000 * (massCleanPost - massEpoxiedPre) / preSA / yearsDeployed,
    deltaVolume_mm3cm2y    = 1000 * (volumeBlockPost + accretion - volumeBlockPre) / preSA / yearsDeployed,
    deltaDensity_mgcm3cm2y = 1000 * (densityCTPost - densityCTPre) / preSA / yearsDeployed,
    macroboring_mm3cm2y    = 1000 * macroboring / preSA / yearsDeployed,
    accretion_mm3cm2y      = 1000 * accretion   / preSA / yearsDeployed,
    grazing_mm3cm2y        = 1000 * grazing     / preSA / yearsDeployed
  )

vars <- c(
  "deltaMass_mgcm2y",
  "deltaMass_kgm2y",
  "deltaVolume_mm3cm2y",
  "deltaDensity_mgcm3cm2y",
  "macroboring_mm3cm2y",
  "accretion_mm3cm2y",
  "grazing_mm3cm2y"
)


```

### Graph response distributions, remove bar outliers over 3sd
```{r dealWithOutliers, echo=FALSE}
# =========================
# (2) Multipanel histograms with ±3 SD highlighting
# =========================
long_df <- barData %>% ###JTM Note: changed from barData1 to barData to run outlier identification on NA filtered dataset
  pivot_longer(cols = all_of(vars), names_to = "variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(variable) %>%
  mutate(
    mu  = mean(value, na.rm = TRUE),
    sdv = sd(value,   na.rm = TRUE),
    lower = mu - 3 * sdv,   # <-- CHANGED
    upper = mu + 3 * sdv,   # <-- CHANGED
    outlier = value < lower | value > upper
  ) %>%
  ungroup()

ggplot(long_df, aes(x = value, fill = outlier)) +
  geom_histogram(bins = 30, color = "black") +
  scale_fill_manual(values = c(`FALSE` = "grey80", `TRUE` = "red"),
                    name = "Outside ±3 SD") +  # <-- CHANGED
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  labs(
    title = "QA/QC Histograms of BAR Variables",
    x = NULL, y = "Frequency"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")

# =========================
# (2A) Table of outliers (values beyond ±3 SD)
# =========================
outlier_table <- barData %>% ###JTM Note: changed from barData1 to barData to run outlier identification on NA filtered dataset
  mutate(tag = .data[["tagNum"]]) %>% #Added this line based on above note
  pivot_longer(cols = all_of(vars), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  mutate(
    mu  = mean(value, na.rm = TRUE),
    sdv = sd(value, na.rm = TRUE),
    lower = mu - 3 * sdv,   # <-- CHANGED
    upper = mu + 3 * sdv,   # <-- CHANGED
    outlier_value = if_else(value < lower | value > upper, as.character(value), "")
  ) %>%
  ungroup() %>%
  select(tag, variable, outlier_value) %>%
  pivot_wider(names_from = variable, values_from = outlier_value) %>%
  filter(if_any(all_of(vars), ~ .x != "")) %>%
  arrange(tag)

pretty_names_out <- names(outlier_table)
pretty_names_out[pretty_names_out == "tag"] <- "Bar tag"

outlier_table %>%
  mutate(across(all_of(vars), ~ ifelse(. == "", "&nbsp;", .))) %>%
  kbl(
    escape    = FALSE,
    col.names = pretty_names_out,
    booktabs  = TRUE,
    align     = "l",
    caption   = "Values Outside ±3 SD (by variable)"  # <-- CHANGED
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "condensed", "hover"),
    full_width        = FALSE,
    position          = "left",
    fixed_thead       = TRUE
  ) %>%
  scroll_box(height = "420px", width = "100%")


cat("Bars to remove due to ±3 SD outliers:", dplyr::n_distinct(outlier_table$tag), "\n")

# filter out outliers
barData <- barData %>%
  filter(!tagNum %in% outlier_table$tag)
```

### Graph deployment duration
```{r deploymentHistogram, include=FALSE}
# =========================
# Histogram of duration with ±3 SD highlighting
# =========================
durationDistribution <- ggplot(barData, aes(x = duration)) +
  geom_histogram(bins = 30, color = "black", fill = "grey80") +
  labs(
    title = "Distribution of Duration",
    x = "Duration",
    y = "Frequency"
  ) +
  theme_minimal(base_size = 12)
durationDistribution

## Regression of deployment distribution
lm_model_deltaMassVsDays <- lm(deltaMass_mgcm2y ~ daysDeployed, data = barData)
r_squared_deltaMassVsDays <- summary(lm_model_deltaMassVsDays)$r.squared
plotDeltaMassVsDays <- ggplot(barData, aes(x = daysDeployed, y = deltaMass_mgcm2y)) +
  geom_point(aes(color = factor(loc1.2)))+  # Scatterplot
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_minimal() +
  scale_color_manual(values = loc_colors) +
  geom_text(data = subset(barData, daysDeployed > 2500), aes(label = tagNum), hjust = -0.1, vjust = -0.5, size = 2)+# Add tag numbers here where days deployed is greater than 2500
  geom_text(data = subset(barData, deltaMass_mgcm2y > 100), aes(label = tagNum), hjust = -0.1, vjust = -0.5, size = 2)+# Add tag numbers here where delta mass is greater than 100
  geom_text(x = 2000, y = 100, label = paste("R^2 =", round(r_squared_deltaMassVsDays, 5)), size = 3)
#Plot of deltaVolume_mm3cm2y vs. daysDeployed:
lm_model_deltaVolumeVsDays <- lm(deltaVolume_mm3cm2y ~ daysDeployed, data = barData)
r_squared_deltaVolumeVsDays <- summary(lm_model_deltaVolumeVsDays)$r.squared
plotDeltaVolumeVsDays <- ggplot(barData, aes(x = daysDeployed, y = deltaVolume_mm3cm2y)) +
  geom_point(aes(color = factor(loc1.2))) +  # Scatterplot
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_minimal()+
  scale_color_manual(values = loc_colors) +
  geom_text(x = 2000, y = -30, label = paste("R^2 =", round(r_squared_deltaVolumeVsDays, 5)), size = 3)
#Plot of deltaDensity_mgcm3cm2y vs. daysDeployed:
lm_model_deltaDensityVsDays <- lm(deltaDensity_mgcm3cm2y ~ daysDeployed, data = barData)
r_squared_deltaDensityVsDays <- summary(lm_model_deltaDensityVsDays)$r.squared
plotDeltaDensityVsDays <- ggplot(barData, aes(x = daysDeployed, y = deltaDensity_mgcm3cm2y)) +
  geom_point(aes(color = factor(loc1.2))) +  # Scatterplot
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_minimal()+
  scale_color_manual(values = loc_colors) +
  geom_text(x = 2000, y = -5, label = paste("R^2 =", round(r_squared_deltaDensityVsDays, 5)), size = 3)
#Plot of macroboring_mm3cm2y vs. daysDeployed:
lm_model_macroboringVsDays <- lm(macroboring_mm3cm2y ~ daysDeployed, data = barData)
r_squared_macroboringVsDays <- summary(lm_model_macroboringVsDays)$r.squared
plotMacroboringVsDays <- ggplot(barData, aes(x = daysDeployed, y = macroboring_mm3cm2y)) +
  geom_point(aes(color = factor(loc1.2))) +  # Scatterplot
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_minimal()+
  scale_color_manual(values = loc_colors) +
  geom_text(x = 2000, y = -30, label = paste("R^2 =", round(r_squared_macroboringVsDays, 5)), size = 3)
#Plot of grazing_mm3cm2y vs. daysDeployed:
lm_model_grazingVsDays <- lm(grazing_mm3cm2y ~ daysDeployed, data = barData)
r_squared_grazingVsDays <- summary(lm_model_grazingVsDays)$r.squared
plotGrazingVsDays <- ggplot(barData, aes(x = daysDeployed, y = grazing_mm3cm2y)) +
  geom_point(aes(color = factor(loc1.2))) +  # Scatterplot
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_minimal()+
  scale_color_manual(values = loc_colors) +
  geom_text(x = 2000, y = -75, label = paste("R^2 =", round(r_squared_grazingVsDays, 2)), size = 3)
#Plot of accretion_mm3cm2y vs. daysDeployed:
lm_model_accretionVsDays <- lm(accretion_mm3cm2y ~ daysDeployed, data = barData)
r_squared_accretionVsDays <- summary(lm_model_accretionVsDays)$r.squared
plotAccretionVsDays <- ggplot(barData, aes(x = daysDeployed, y = accretion_mm3cm2y)) +
  geom_point(aes(color = factor(loc1.2))) +  # Scatterplot
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  theme_minimal()+
  scale_color_manual(values = loc_colors) +
  geom_text(x = 2000, y = 100, label = paste("R^2 =", round(r_squared_accretionVsDays, 2)), size = 3)
ggarrange(plotDeltaMassVsDays,plotDeltaVolumeVsDays,plotDeltaDensityVsDays,plotMacroboringVsDays,plotGrazingVsDays,plotAccretionVsDays, ncol = 3, nrow = 2, common.legend = TRUE, legend = "bottom")
```

### Clean data
Remove data that is
1. Not successfully recovered
2. Not plob or mcav
3. Tag 1022
4. Data from Maug seep
5. Grazing data that is positive
6. Deployed for more than 6 years or less than 2 years
7. Depth is less than 10m
```{r cleanData, echo=FALSE}
barData <- barData %>%
  filter(fate == "SUCCESSFULLY RECOVERED",
         species != "iso",
         tagNum != 1022,
         loc3 != "maug-innerMCO2",
         loc3 != "maug-innerHCO2",
         grazing_mm3cm2y < 0 | is.na(grazing_mm3cm2y),
         duration < 6,
         duration > 2,
         depth >= 10)
```

### Remove vars with low replication
``` {r removeLowReplicationVars, echo=FALSE}
min_reps <- 3
# ---- helper: mask one variable based on loc3 replication ----
mask_low_rep_one <- function(df, var, min_reps) {
  var_sym <- rlang::sym(var)

  df %>%
    dplyr::group_by(loc3) %>%
    dplyr::mutate(
      .n_non_na = sum(!is.na(!!var_sym)),
      !!var_sym := dplyr::if_else(
        .n_non_na < min_reps,
        NA_real_,
        as.numeric(!!var_sym)
      )
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(-.n_non_na)
}

# ---- apply to each variable ----
for (v in vars) {
  barData <- mask_low_rep_one(barData, v, min_reps)
}

# ---- summary table ----
mask_summary <- purrr::map_dfr(vars, function(v) {
  var_sym <- rlang::sym(v)

  barData %>%
    dplyr::group_by(loc3) %>%
    dplyr::summarise(
      n_non_na = sum(!is.na(!!var_sym)),
      .groups = "drop"
    ) %>%
    dplyr::summarise(
      var = v,
      n_locs_total = dplyr::n(),
      n_locs_below_min = sum(n_non_na < min_reps)
    )
})

cat(sprintf(
  "### Values masked (set to NA) where loc3 has < %d non-NA measurements (per variable)\n",
  min_reps
))

if (requireNamespace("knitr", quietly = TRUE)) {
  knitr::kable(mask_summary)
} else {
  print(mask_summary)
}
```

### Remove rows that have NA's for all vars
```{r dropAllNAvars, echo=FALSE}
barData <- barData %>%
  dplyr::filter(
    !if_all(dplyr::all_of(vars), ~ is.na(.x))
  )
```

### Check binds across datasets
```{r checkBinds, echo=FALSE}
# ===== Function (general 3-way compare) =====
compare_loc <- function(df_bar, df_parrot, df_env, loc_col) {
  join_col <- rlang::as_name(rlang::ensym(loc_col))

  full_join(
    df_bar    %>% distinct({{loc_col}}) %>% mutate(in_BAR = TRUE),
    df_parrot %>% distinct({{loc_col}}) %>% mutate(in_PARROT = TRUE),
    by = join_col
  ) %>%
    full_join(
      df_env %>% distinct({{loc_col}}) %>% mutate(in_ENV = TRUE),
      by = join_col
    ) %>%
    # Any missing membership flags become FALSE
    mutate(across(starts_with("in_"), ~ tidyr::replace_na(., FALSE))) %>%
    mutate(
      match_status = case_when(
        in_BAR & in_PARROT & in_ENV                  ~ "MATCH",
        in_BAR & in_PARROT & !in_ENV                 ~ "BAR & PARROT only",
        in_BAR & in_ENV    & !in_PARROT              ~ "BAR & ENV only",
        in_PARROT & in_ENV & !in_BAR                 ~ "PARROT & ENV only",
        in_BAR    & !in_PARROT & !in_ENV            ~ "ONLY in BAR",
        in_PARROT & !in_BAR    & !in_ENV            ~ "ONLY in PARROT",
        in_ENV    & !in_BAR    & !in_PARROT         ~ "ONLY in ENV",
        TRUE                                          ~ "UNKNOWN"
      )
    ) %>%
    arrange({{loc_col}})
}

# ===== Function (original 2-way BAR vs PARROT only) =====
compare_loc_2way <- function(df_bar, df_parrot, loc_col) {
  join_col <- rlang::as_name(rlang::ensym(loc_col))

  full_join(
    df_bar    %>% distinct({{loc_col}}) %>% mutate(in_BAR = TRUE),
    df_parrot %>% distinct({{loc_col}}) %>% mutate(in_PARROT = TRUE),
    by = join_col
  ) %>%
    mutate(across(starts_with("in_"), ~ tidyr::replace_na(., FALSE))) %>%
    mutate(
      match_status = case_when(
        in_BAR & in_PARROT  ~ "MATCH",
        in_BAR & !in_PARROT ~ "ONLY in BAR",
        !in_BAR & in_PARROT ~ "ONLY in PARROT",
        TRUE                ~ "UNKNOWN"
      )
    ) %>%
    arrange({{loc_col}})
}

# ===== Run comparisons =====
loc1_compare <- compare_loc(barData, parrotData, environmentalData, loc1)
loc2_compare <- compare_loc(barData, parrotData, environmentalData, loc2)
loc3_compare <- compare_loc(barData, parrotData, environmentalData, loc3)  # <-- now 3-way

# ===== View results =====
kable(loc1_compare, caption = "loc1 Comparison: BAR vs PARROT vs ENV")
kable(loc2_compare, caption = "loc2 Comparison: BAR vs PARROT vs ENV")
kable(loc3_compare, caption = "loc3 Comparison: BAR vs PARROT vs ENV")
```

### Removes Env and Parrot data if parrot and env only, or only in env, or only parrot
```{r removeEnvWithoutBAR_loc3, echo=FALSE}
# loc3s that should NOT exist outside BAR
bad_loc3 <- loc3_compare %>%
  dplyr::filter(match_status %in% c("PARROT & ENV only", "ONLY in ENV", "ONLY in PARROT")) %>%
  dplyr::pull(loc3) %>%
  unique()

cat("Removing loc3 missing from BAR:", length(bad_loc3), "\n")

# ---- filter parrot + env ONLY ----
parrotData <- parrotData %>%
  dplyr::filter(!(loc3 %in% bad_loc3))

environmentalData <- environmentalData %>%
  dplyr::filter(!(loc3 %in% bad_loc3))
```

### Check binds across datasets again after removal
```{r checkBindsAgain, echo=FALSE}
# ===== Function (general 3-way compare) =====
compare_loc <- function(df_bar, df_parrot, df_env, loc_col) {
  join_col <- rlang::as_name(rlang::ensym(loc_col))

  full_join(
    df_bar    %>% distinct({{loc_col}}) %>% mutate(in_BAR = TRUE),
    df_parrot %>% distinct({{loc_col}}) %>% mutate(in_PARROT = TRUE),
    by = join_col
  ) %>%
    full_join(
      df_env %>% distinct({{loc_col}}) %>% mutate(in_ENV = TRUE),
      by = join_col
    ) %>%
    # Any missing membership flags become FALSE
    mutate(across(starts_with("in_"), ~ tidyr::replace_na(., FALSE))) %>%
    mutate(
      match_status = case_when(
        in_BAR & in_PARROT & in_ENV                  ~ "MATCH",
        in_BAR & in_PARROT & !in_ENV                 ~ "BAR & PARROT only",
        in_BAR & in_ENV    & !in_PARROT              ~ "BAR & ENV only",
        in_PARROT & in_ENV & !in_BAR                 ~ "PARROT & ENV only",
        in_BAR    & !in_PARROT & !in_ENV            ~ "ONLY in BAR",
        in_PARROT & !in_BAR    & !in_ENV            ~ "ONLY in PARROT",
        in_ENV    & !in_BAR    & !in_PARROT         ~ "ONLY in ENV",
        TRUE                                          ~ "UNKNOWN"
      )
    ) %>%
    arrange({{loc_col}})
}

# ===== Function (original 2-way BAR vs PARROT only) =====
compare_loc_2way <- function(df_bar, df_parrot, loc_col) {
  join_col <- rlang::as_name(rlang::ensym(loc_col))

  full_join(
    df_bar    %>% distinct({{loc_col}}) %>% mutate(in_BAR = TRUE),
    df_parrot %>% distinct({{loc_col}}) %>% mutate(in_PARROT = TRUE),
    by = join_col
  ) %>%
    mutate(across(starts_with("in_"), ~ tidyr::replace_na(., FALSE))) %>%
    mutate(
      match_status = case_when(
        in_BAR & in_PARROT  ~ "MATCH",
        in_BAR & !in_PARROT ~ "ONLY in BAR",
        !in_BAR & in_PARROT ~ "ONLY in PARROT",
        TRUE                ~ "UNKNOWN"
      )
    ) %>%
    arrange({{loc_col}})
}

# ===== Run comparisons =====
loc1_compare <- compare_loc(barData, parrotData, environmentalData, loc1)
loc2_compare <- compare_loc(barData, parrotData, environmentalData, loc2)
loc3_compare <- compare_loc(barData, parrotData, environmentalData, loc3)  # <-- now 3-way

# ===== View results =====
kable(loc1_compare, caption = "loc1 Comparison: BAR vs PARROT vs ENV")
kable(loc2_compare, caption = "loc2 Comparison: BAR vs PARROT vs ENV")
kable(loc3_compare, caption = "loc3 Comparison: BAR vs PARROT vs ENV")
```

### Count numbers of replicates of each variable by loc2 and loc3
```{r counts_by_loc2_loc3, echo=FALSE, message=FALSE, warning=FALSE}
# Safety check (prevents the pivot_longer-style error you hit earlier)
stopifnot(is.character(vars))
# Counts by loc2
count_loc2 <- barData %>%
  group_by(loc2) %>%
  summarise(
    across(all_of(vars), ~ sum(!is.na(.x)), .names = "{.col}"),
    n_rows = n(),
    .groups = "drop"
  ) %>%
  relocate(n_rows, .after = loc2) %>%
  arrange(desc(n_rows)) %>%
  bind_rows(
    summarise(
      .,
      loc2   = "TOTAL",
      n_rows = sum(n_rows),
      across(all_of(vars), sum)
    )
  )

count_loc2 %>%
  kbl(
    caption = "Non-missing counts by loc2 (per variable)"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  scroll_box(
    height = "450px",
    width  = "100%"
  )

# Counts by loc3
count_loc3 <- barData %>%
  group_by(loc3) %>%
  summarise(
    across(all_of(vars), ~ sum(!is.na(.x)), .names = "{.col}"),
    n_rows = n(),
    .groups = "drop"
  ) %>%
  relocate(n_rows, .after = loc3) %>%
  arrange(desc(n_rows)) %>%
  bind_rows(
    summarise(
      .,
      loc3   = "TOTAL",
      n_rows = sum(n_rows),
      across(all_of(vars), sum)
    )
  )

count_loc3 %>%
  kbl(
    caption = "Non-missing counts by loc3 (per variable)"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  scroll_box(
    height = "450px",
    width  = "100%"
  )
```



# Data exploration
## General
### Map of BAR, parrot surveys, and environmental sites
```{r, echo=FALSE}
barSiteMap <- leaflet() %>%
  addProviderTiles(providers$CartoDB.PositronNoLabels) %>%

  # ===== BAR DATA =====
  addCircleMarkers(
    data = barData,
    lng = ~longitude, lat = ~latitude,
    radius = 2, color = "blue",
    label = ~paste0(loc3),
    labelOptions = labelOptions(
      noHide = TRUE,
      direction = "top",
      textsize = "12px",
      offset = c(0, -4)
    )
  ) %>%

  # ===== PARROTFISH DATA =====
  addCircleMarkers(
    data = parrotData,
    lng = ~longitudeFishSurvey, lat = ~latitudeFishSurvey,
    radius = 1, color = "red",
    popup = ~paste0(loc3,",",siteFishSurvey)   
  )   %>%

  # ===== ENVIRONMENTAL DATA (COMMENTED OUT) =====

  addCircleMarkers(
    data = environmentalData,
    lng = ~lon, lat = ~lat,
    radius = 4, color = "green",
    popup = ~paste0(loc3,",",lat,",",lon)   
  ) %>%

  addCircleMarkers(
    data = environmentalData_raw,
    lng = ~lon, lat = ~lat,
    radius = .5, color = "orange",
    popup = ~paste0(lat,",",lon)   
  )


barSiteMap
```

### Timeline of BAR data, parrot surveys, and environmental data
```{r, echo=FALSE}
# -----------------------------
# 0) Appearance knobs (tweak as desired)
# -----------------------------
tick_h        <- 0.18  # half-height of deploy/recover vertical ticks
sst_y_offset  <- 0.25  # SST dots sit this far below the site line
chl_y_offset  <- 0.35  # chlorophyll dots sit slightly further down
sst_col       <- "mean_Sea_Surface_Temperature_CRW_daily_01mo"  # <— your SST column

# Helper: pick a likely chlorophyll column if you don't want to hardcode it
pick_chl_col <- function(df) {
  cols <- names(df)
  cand <- cols[grepl("chlor|chla|chl_a|chlorophyll", cols, ignore.case = TRUE)]
  # prefer "mean" if present
  cand <- c(cand[grepl("mean", cand, ignore.case = TRUE)],
            setdiff(cand, cand[grepl("mean", cand, ignore.case = TRUE)]))
  if (length(cand)) cand[1] else NA_character_
}

# If you prefer to hardcode a chlorophyll column, set it here (uncomment & edit):
# chl_col <- "mean_Chlorophyll_MODIS_daily_01mo"
chl_col <- pick_chl_col(environmentalData)

# -----------------------------
# 1) Timeline data (deploy/recover)
# -----------------------------
timeline_df <- barData %>%
  mutate(
    deploy  = ymd(dateDeployment),
    recover = ymd(dateRecovery)
  ) %>%
  filter(!is.na(loc1.2), !is.na(deploy), !is.na(recover)) %>%
  mutate(
    start = pmin(deploy, recover, na.rm = TRUE),
    end   = pmax(deploy, recover, na.rm = TRUE)
  ) %>%
  group_by(loc1.2) %>%
  arrange(start, .by_group = TRUE) %>%
  ungroup()

# order sites by first deployment and freeze that order
site_levels <- timeline_df %>%
  group_by(loc1.2) %>%
  summarise(first_start = min(start, na.rm = TRUE), .groups = "drop") %>%
  arrange(first_start) %>%
  pull(loc1.2)

timeline_df <- timeline_df %>%
  mutate(
    loc1.2 = factor(loc1.2, levels = site_levels),
    y_num  = as.numeric(loc1.2)
  )

# -----------------------------
# 2) Parrot survey data (black points, mid-year)
# -----------------------------
survey_df <- parrotData %>%
  filter(!is.na(loc1.2), !is.na(yearFishSurvey)) %>%
  mutate(
    year_date = ymd(paste0(as.character(yearFishSurvey), "-07-01")),
    loc1.2    = factor(loc1.2, levels = site_levels),
    y_num     = as.numeric(loc1.2)
  ) %>%
  filter(!is.na(year_date), !is.na(y_num))

# -----------------------------
# 3) Environmental SST (light red dots) — date is DMY
# -----------------------------
sst_df <- environmentalData %>%
  mutate(
    date   = dmy(date),             # <-- your environmentalData date format
    sstval = .data[[sst_col]],
    loc1.2 = factor(loc1.2, levels = site_levels),
    y_num  = as.numeric(loc1.2)
  ) %>%
  filter(!is.na(loc1.2), !is.na(date), !is.na(sstval), !is.na(y_num))

# -----------------------------
# 4) Environmental Chlorophyll (light green dots) — date is DMY
# -----------------------------
chl_df <- environmentalData %>%
  mutate(
    date   = dmy(date),
    chlval = if (!is.na(chl_col)) .data[[chl_col]] else NA_real_,
    loc1.2 = factor(loc1.2, levels = site_levels),
    y_num  = as.numeric(loc1.2)
  ) %>%
  filter(!is.na(loc1.2), !is.na(date), !is.na(chlval), !is.na(y_num))

# -----------------------------
# 5) Plot
# -----------------------------
p_timeline <- ggplot() +
  # Grey bar: deployment → recovery
  geom_segment(
    data = timeline_df,
    aes(x = start, xend = end, y = y_num, yend = y_num),
    linewidth = 3, lineend = "round", color = "grey50"
  ) +
  # Deployment tick (green)
  geom_segment(
    data = timeline_df,
    aes(x = start, xend = start, y = y_num - tick_h, yend = y_num + tick_h),
    linewidth = 0.9, color = "forestgreen"
  ) +
  # Recovery tick (red)
  geom_segment(
    data = timeline_df,
    aes(x = end, xend = end, y = y_num - tick_h, yend = y_num + tick_h),
    linewidth = 0.9, color = "firebrick"
  ) +
  # Parrot survey years (black)
  geom_point(
    data = survey_df,
    aes(x = year_date, y = y_num),
    size = 2.2, color = "black", alpha = 0.9
  ) +
  # SST dots (light red) with legend
  geom_point(
    data = sst_df,
    aes(x = date, y = y_num - sst_y_offset, color = "Mean SST (01mo)"),
    size = 0.9, alpha = 0.8
  ) +
  # Chlorophyll dots (light green) with legend
  geom_point(
    data = chl_df,
    aes(x = date, y = y_num - chl_y_offset, color = "Chlorophyll"),
    size = 0.7, alpha = 0.8
  ) +
  # Legend (SST + Chlor) at bottom
  scale_color_manual(
    name   = NULL,
    values = c("Mean SST (01mo)" = "tomato", "Chlorophyll" = "palegreen3"),
    breaks = c("Mean SST (01mo)", "Chlorophyll")
  ) +
  guides(
    color = guide_legend(
      override.aes = list(size = 3, alpha = 1, shape = 16)
    )
  ) +
  # Axes & theme
  scale_x_date(name = NULL, date_breaks = "6 months", date_labels = "%Y-%m") +
  scale_y_continuous(
    name = NULL,
    breaks = seq_along(site_levels),
    labels = site_levels
  ) +
  ggtitle("Deployments (green), Recoveries (red), Surveys (black), SST (light red), Chlorophyll (light green)") +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "bottom"
  )

p_timeline
```



## BAR data
### Summary of locations and durations
```{r barDeploymentSummaryData, include=FALSE}
nLoc1 <-length(unique(barData$loc1))
listLoc1 <-unique(barData$loc1)
nLoc2 <-length(unique(barData$loc2))
listLoc2 <-unique(barData$loc2)
nLoc3 <-length(unique(barData$loc3))
minDeployment <- min(barData$dateDeployment)
maxDeployment <- max(barData$dateDeployment)
minRecovery <- min(barData$dateRecovery)
maxRecovery <- max(barData$dateRecovery)
minDuration <- min(barData$daysDeployed)
maxDuration <- max(barData$daysDeployed)
meanDuration <- mean(barData$daysDeployed)
```

Sites are divided at three levels across 2 basins. The highest level (loc1) includes `r nLoc1` large regions including:  
<small>`r listLoc1`</small>  
  
Within these larger regions are `r nLoc2` sublevels (loc2) that are generally inclusive of an island including:  
<small>`r listLoc2`</small>  
  
Most islands or loc2 level includes multiple sites (loc3) within it, totaling `r nLoc3` loc3 levels across both basins
  
Deployment dates ranged from `r minDeployment` to `r maxDeployment` and recovery dates ranged from `r minRecovery` to `r maxRecovery`. The deployment duration ranged from `r minDuration` to `r maxDuration` days and was on `r meanDuration` days.

### Calculation of stats for bar alteration by loc2
```{r barStatCalculations_loc2, echo=FALSE}
barStatsLoc2 <- barData %>%
  group_by(basin, loc1, loc2, loc1.2) %>%
  summarise(
    # duration
    maxDays = max(daysDeployed, na.rm = TRUE),
    minDays = min(daysDeployed, na.rm = TRUE),
    n       = dplyr::n(),

    # depth
    maxDepth  = max(depth, na.rm = TRUE),
    minDepth  = min(depth, na.rm = TRUE),
    meanDepth = mean(depth, na.rm = TRUE),

    # deltaMass_mgcm2y
    meanDeltaMass_mgcm2y  = mean(deltaMass_mgcm2y, na.rm = TRUE),
    countDeltaMass_mgcm2y = sum(!is.na(deltaMass_mgcm2y)),
    sdDeltaMass_mgcm2y    = sd(deltaMass_mgcm2y, na.rm = TRUE),
    seDeltaMass_mgcm2y    = sdDeltaMass_mgcm2y / sqrt(countDeltaMass_mgcm2y),

    # deltaMass_kgm2y
    meanDeltaMass_kgm2y  = mean(deltaMass_kgm2y, na.rm = TRUE),
    countDeltaMass_kgm2y = sum(!is.na(deltaMass_kgm2y)),
    sdDeltaMass_kgm2y    = sd(deltaMass_kgm2y, na.rm = TRUE),
    seDeltaMass_kgm2y    = sdDeltaMass_kgm2y / sqrt(countDeltaMass_kgm2y),    
      
    # deltaVolume
    meanDeltaVolume  = mean(deltaVolume_mm3cm2y, na.rm = TRUE),
    countDeltaVolume = sum(!is.na(deltaVolume_mm3cm2y)),
    sdDeltaVolume    = sd(deltaVolume_mm3cm2y, na.rm = TRUE),
    seDeltaVolume    = sdDeltaVolume / sqrt(countDeltaVolume),
  
    # deltaDensity
    meanDeltaDensity  = mean(deltaDensity_mgcm3cm2y, na.rm = TRUE),
    countDeltaDensity = sum(!is.na(deltaDensity_mgcm3cm2y)),
    sdDeltaDensity    = sd(deltaDensity_mgcm3cm2y, na.rm = TRUE),
    seDeltaDensity    = sdDeltaDensity / sqrt(countDeltaDensity),
  
    # macroboring
    meanMacroboring  = mean(macroboring_mm3cm2y, na.rm = TRUE),
    countMacroboring = sum(!is.na(macroboring_mm3cm2y)),
    sdMacroboring    = sd(macroboring_mm3cm2y, na.rm = TRUE),
    seMacroboring    = sdMacroboring / sqrt(countMacroboring),
  
    # accretion
    meanAccretion  = mean(accretion_mm3cm2y, na.rm = TRUE),
    countAccretion = sum(!is.na(accretion_mm3cm2y)),
    sdAccretion    = sd(accretion_mm3cm2y, na.rm = TRUE),
    seAccretion    = sdAccretion / sqrt(countAccretion),
    
    # grazing
    meanGrazing  = mean(grazing_mm3cm2y, na.rm = TRUE),
    countGrazing = sum(!is.na(grazing_mm3cm2y)),
    sdGrazing    = sd(grazing_mm3cm2y, na.rm = TRUE),
    seGrazing    = sdGrazing / sqrt(countGrazing),

  
    .groups = "drop"
  )

kbl(barStatsLoc2, align = "l") %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  add_header_above(c(
    "location"     = 4,
    "duration"     = 3,  # maxDays, minDays, n
    "depth"        = 3,
    "deltaMass_mgcm2yr"= 4,
    "deltaMass_kgm2y"= 4,
    "deltaVolume"  = 4,
    "deltaDensity" = 4,
    "macroboring"  = 4,
    "accretion"    = 4,
    "grazing"      = 4
  )) %>%
  scroll_box(width = "1000px", height = "500px")
```
### Calculation of stats for bar alteration by loc3
```{r barStatCalculations_loc3, echo=FALSE}
barStatsLoc3 <- barData %>%
  filter(!is.na(grazing_mm3cm2y)) %>%
  #group_by(basin, loc1, loc2, loc1.2, loc3, latitude, longitude) %>% ##Note: grouping with lat/long is problematic since some Loc3 sites are marginally different lat/long
  group_by(basin, loc1, loc2, loc1.2, loc3) %>%
  summarise(
    # duration
    maxDays = max(daysDeployed, na.rm = TRUE),
    minDays = min(daysDeployed, na.rm = TRUE),

    # depth
    maxDepth  = max(depth, na.rm = TRUE),
    minDepth  = min(depth, na.rm = TRUE),
    meanDepth = mean(depth, na.rm = TRUE),

    # deltaMass_mgcm2y
    meanDeltaMass_mgcm2y  = mean(deltaMass_mgcm2y, na.rm = TRUE),
    countDeltaMass_mgcm2y = sum(!is.na(deltaMass_mgcm2y)),
    sdDeltaMass_mgcm2y    = sd(deltaMass_mgcm2y, na.rm = TRUE),
    seDeltaMass_mgcm2y    = sdDeltaMass_mgcm2y / sqrt(countDeltaMass_mgcm2y),

    # deltaMass_kgm2y
    meanDeltaMass_kgm2y  = mean(deltaMass_kgm2y, na.rm = TRUE),
    countDeltaMass_kgm2y = sum(!is.na(deltaMass_kgm2y)),
    sdDeltaMass_kgm2y    = sd(deltaMass_kgm2y, na.rm = TRUE),
    seDeltaMass_kgm2y    = sdDeltaMass_kgm2y / sqrt(countDeltaMass_kgm2y),    
      
    # deltaVolume
    meanDeltaVolume  = mean(deltaVolume_mm3cm2y, na.rm = TRUE),
    countDeltaVolume = sum(!is.na(deltaVolume_mm3cm2y)),
    sdDeltaVolume    = sd(deltaVolume_mm3cm2y, na.rm = TRUE),
    seDeltaVolume    = sdDeltaVolume / sqrt(countDeltaVolume),
  
    # deltaDensity
    meanDeltaDensity  = mean(deltaDensity_mgcm3cm2y, na.rm = TRUE),
    countDeltaDensity = sum(!is.na(deltaDensity_mgcm3cm2y)),
    sdDeltaDensity    = sd(deltaDensity_mgcm3cm2y, na.rm = TRUE),
    seDeltaDensity    = sdDeltaDensity / sqrt(countDeltaDensity),
  
    # macroboring
    meanMacroboring  = mean(macroboring_mm3cm2y, na.rm = TRUE),
    countMacroboring = sum(!is.na(macroboring_mm3cm2y)),
    sdMacroboring    = sd(macroboring_mm3cm2y, na.rm = TRUE),
    seMacroboring    = sdMacroboring / sqrt(countMacroboring),
  
    # accretion
    meanAccretion  = mean(accretion_mm3cm2y, na.rm = TRUE),
    countAccretion = sum(!is.na(accretion_mm3cm2y)),
    sdAccretion    = sd(accretion_mm3cm2y, na.rm = TRUE),
    seAccretion    = sdAccretion / sqrt(countAccretion),
    
    # grazing
    meanGrazing  = mean(grazing_mm3cm2y, na.rm = TRUE),
    countGrazing = sum(!is.na(grazing_mm3cm2y)),
    sdGrazing    = sd(grazing_mm3cm2y, na.rm = TRUE),
    seGrazing    = sdGrazing / sqrt(countGrazing),

  
    .groups = "drop"
  )

kbl(barStatsLoc3, align = "l") %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  add_header_above(c(
    "location"     = 4,
    "duration"     = 3,  # maxDays, minDays, n
    "depth"        = 3,
    "deltaMass_mgcm2yr"= 4,
    "deltaMass_kgm2y"= 4,
    "deltaVolume"  = 4,
    "deltaDensity" = 4,
    "macroboring"  = 4,
    "accretion"    = 4,
    "grazing"      = 4
  )) %>%
  scroll_box(width = "1000px", height = "500px")
```


## Parrot fish community
### Summary
```{r, include=FALSE}
nParrotSurveys <- nrow(parrotData)
nParrotLoc1 <-length(unique(parrotData$loc1))
listParrotLoc1 <-unique(parrotData$loc1)
nParrotLoc2 <-length(unique(parrotData$loc2))
listParrotLoc2 <-unique(parrotData$loc2)
nParrotLoc3 <-length(unique(parrotData$loc3))
```

There are a total of `r nParrotSurveys` parrot surveys included in this study. Sites are divided at three levels across 2 basins. The highest level (loc1) includes `r nParrotLoc1` large regions including:  
<small>`r listParrotLoc1`</small>  

### Unify parrot tables and calculate stats at loc3 level
```{r unifyParrotTablesLoc3, echo=FALSE}

# ---------- Clean & standardize ----------
parrotData_clean <- parrotData %>%
  mutate(
    across(c(basin, loc1, loc2, loc3), ~ .x |> as.character() |> stringr::str_squish()),
    `loc1.2` = if_else(is.na(loc2) | loc2 == "", loc1, paste0(loc1, ".", loc2)) |>
      stringr::str_squish()
  )

# ---------- Community summary (means/SDs + N) ----------
parrot_loc3_summary <- parrotData_clean %>%
  group_by(basin, loc1, loc2, `loc1.2`, loc3) %>%
  summarise(
    # Community
    comm_density_mean    = mean(totalDensity,  na.rm = TRUE),
    comm_density_sd      = sd(totalDensity,    na.rm = TRUE),
    comm_density_n       = sum(!is.na(totalDensity)),
    comm_biomass_mean    = mean(totalBiomass,  na.rm = TRUE),
    comm_biomass_sd      = sd(totalBiomass,    na.rm = TRUE),
    comm_biomass_n       = sum(!is.na(totalBiomass)),
    comm_bioerosion_mean = mean(totalBioerosion, na.rm = TRUE),
    comm_bioerosion_sd   = sd(totalBioerosion,   na.rm = TRUE),
    comm_bioerosion_n    = sum(!is.na(totalBioerosion)),

    # Scraper
    scraper_density_mean    = mean(scraperDensity,  na.rm = TRUE),
    scraper_density_sd      = sd(scraperDensity,    na.rm = TRUE),
    scraper_density_n       = sum(!is.na(scraperDensity)),
    scraper_biomass_mean    = mean(scraperBiomass,  na.rm = TRUE),
    scraper_biomass_sd      = sd(scraperBiomass,    na.rm = TRUE),
    scraper_biomass_n       = sum(!is.na(scraperBiomass)),
    scraper_bioerosion_mean = mean(scraperBioerosion, na.rm = TRUE),
    scraper_bioerosion_sd   = sd(scraperBioerosion,   na.rm = TRUE),
    scraper_bioerosion_n    = sum(!is.na(scraperBioerosion)),

    # Excavator
    excavator_density_mean    = mean(excavatorDensity,  na.rm = TRUE),
    excavator_density_sd      = sd(excavatorDensity,    na.rm = TRUE),
    excavator_density_n       = sum(!is.na(excavatorDensity)),
    excavator_biomass_mean    = mean(excavatorBiomass,  na.rm = TRUE),
    excavator_biomass_sd      = sd(excavatorBiomass,    na.rm = TRUE),
    excavator_biomass_n       = sum(!is.na(excavatorBiomass)),
    excavator_bioerosion_mean = mean(excavatorBioerosion, na.rm = TRUE),
    excavator_bioerosion_sd   = sd(excavatorBioerosion,   na.rm = TRUE),
    excavator_bioerosion_n    = sum(!is.na(excavatorBioerosion)),

    .groups = "drop"
  )

# ---------- Order columns to match headers ----------
id_cols <- c("basin","loc1","loc2","loc1.2","loc3")
metric_order <- c(
  # Community
  "comm_density_mean","comm_density_sd","comm_density_n",
  "comm_biomass_mean","comm_biomass_sd","comm_biomass_n",
  "comm_bioerosion_mean","comm_bioerosion_sd","comm_bioerosion_n",
  # Scrapers
  "scraper_density_mean","scraper_density_sd","scraper_density_n",
  "scraper_biomass_mean","scraper_biomass_sd","scraper_biomass_n",
  "scraper_bioerosion_mean","scraper_bioerosion_sd","scraper_bioerosion_n",
  # Excavators
  "excavator_density_mean","excavator_density_sd","excavator_density_n",
  "excavator_biomass_mean","excavator_biomass_sd","excavator_biomass_n",
  "excavator_bioerosion_mean","excavator_bioerosion_sd","excavator_bioerosion_n"
)

display_tbl <- parrot_loc3_summary %>%
  select(all_of(id_cols), any_of(metric_order)) %>%
  mutate(
    # Density & biomass mean/sd → 1 decimal (all groups)
    across(matches("^(comm|scraper|excavator)_(density|biomass)_(mean|sd)$"), ~ round(., 1)),
    # Bioerosion mean/sd → 2 decimals (all groups)
    across(matches("^(comm|scraper|excavator)_bioerosion_(mean|sd)$"), ~ round(., 2)),
    # N as integers
    across(matches("_n$"), ~ as.integer(.))
  )

# ---------- Alignment (auto) ----------
align_vec <- c(rep("l", length(id_cols)), rep("r", ncol(display_tbl) - length(id_cols)))

# ---------- Headers ----------
col_names <- c(
  "Basin","Loc1","Loc2","Loc1.2","Loc3",
  rep(c("Mean","SD","N"), 9)  # 9 metric blocks (3 groups × 3 metrics)
)

kableExtra::kbl(
  display_tbl,
  booktabs = TRUE,
  align = align_vec,
  col.names = col_names,
  caption = "Parrot community & functional-group summary at loc3 (Mean, SD, N)."
) %>%
  kableExtra::add_header_above(
    c(" " = 5,
      "Density" = 3, "Biomass" = 3, "Bioerosion" = 3,       # Community
      "Density" = 3, "Biomass" = 3, "Bioerosion" = 3,       # Scrapers
      "Density" = 3, "Biomass" = 3, "Bioerosion" = 3),      # Excavators
    bold = FALSE
  ) %>%
  kableExtra::add_header_above(
    c(" " = 5, "Community" = 9, "Scrapers" = 9, "Excavators" = 9)
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 11,
    bootstrap_options = c("striped","hover","condensed","responsive")
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "520px")
```
### Unify parrot tables and calculate stats at loc2 level
```{r unifyParrotTablesLoc2, echo=FALSE}

# ---------- Loc2 summary (means/SDs + N) ----------
parrot_loc2_summary <- parrotData_clean %>%
  group_by(basin, loc1, loc2, `loc1.2`) %>%
  summarise(
    # Community
    comm_density_mean    = mean(totalDensity,  na.rm = TRUE),
    comm_density_sd      = sd(totalDensity,    na.rm = TRUE),
    comm_density_n       = sum(!is.na(totalDensity)),
    comm_biomass_mean    = mean(totalBiomass,  na.rm = TRUE),
    comm_biomass_sd      = sd(totalBiomass,    na.rm = TRUE),
    comm_biomass_n       = sum(!is.na(totalBiomass)),
    comm_bioerosion_mean = mean(totalBioerosion, na.rm = TRUE),
    comm_bioerosion_sd   = sd(totalBioerosion,   na.rm = TRUE),
    comm_bioerosion_n    = sum(!is.na(totalBioerosion)),

    # Scraper
    scraper_density_mean    = mean(scraperDensity,  na.rm = TRUE),
    scraper_density_sd      = sd(scraperDensity,    na.rm = TRUE),
    scraper_density_n       = sum(!is.na(scraperDensity)),
    scraper_biomass_mean    = mean(scraperBiomass,  na.rm = TRUE),
    scraper_biomass_sd      = sd(scraperBiomass,    na.rm = TRUE),
    scraper_biomass_n       = sum(!is.na(scraperBiomass)),
    scraper_bioerosion_mean = mean(scraperBioerosion, na.rm = TRUE),
    scraper_bioerosion_sd   = sd(scraperBioerosion,   na.rm = TRUE),
    scraper_bioerosion_n    = sum(!is.na(scraperBioerosion)),

    # Excavator
    excavator_density_mean    = mean(excavatorDensity,  na.rm = TRUE),
    excavator_density_sd      = sd(excavatorDensity,    na.rm = TRUE),
    excavator_density_n       = sum(!is.na(excavatorDensity)),
    excavator_biomass_mean    = mean(excavatorBiomass,  na.rm = TRUE),
    excavator_biomass_sd      = sd(excavatorBiomass,    na.rm = TRUE),
    excavator_biomass_n       = sum(!is.na(excavatorBiomass)),
    excavator_bioerosion_mean = mean(excavatorBioerosion, na.rm = TRUE),
    excavator_bioerosion_sd   = sd(excavatorBioerosion,   na.rm = TRUE),
    excavator_bioerosion_n    = sum(!is.na(excavatorBioerosion)),

    .groups = "drop"
  )

# ---------- Order columns to match headers ----------
id_cols <- c("basin","loc1","loc2","loc1.2")

metric_order <- c(
  # Community
  "comm_density_mean","comm_density_sd","comm_density_n",
  "comm_biomass_mean","comm_biomass_sd","comm_biomass_n",
  "comm_bioerosion_mean","comm_bioerosion_sd","comm_bioerosion_n",
  # Scrapers
  "scraper_density_mean","scraper_density_sd","scraper_density_n",
  "scraper_biomass_mean","scraper_biomass_sd","scraper_biomass_n",
  "scraper_bioerosion_mean","scraper_bioerosion_sd","scraper_bioerosion_n",
  # Excavators
  "excavator_density_mean","excavator_density_sd","excavator_density_n",
  "excavator_biomass_mean","excavator_biomass_sd","excavator_biomass_n",
  "excavator_bioerosion_mean","excavator_bioerosion_sd","excavator_bioerosion_n"
)

display_tbl <- parrot_loc2_summary %>%
  select(all_of(id_cols), any_of(metric_order)) %>%
  mutate(
    # Density & biomass mean/sd → 1 decimal (all groups)
    across(matches("^(comm|scraper|excavator)_(density|biomass)_(mean|sd)$"), ~ round(., 1)),
    # Bioerosion mean/sd → 2 decimals (all groups)
    across(matches("^(comm|scraper|excavator)_bioerosion_(mean|sd)$"), ~ round(., 2)),
    # N as integers
    across(matches("_n$"), ~ as.integer(.))
  )

# ---------- Alignment (auto) ----------
align_vec <- c(rep("l", length(id_cols)), rep("r", ncol(display_tbl) - length(id_cols)))

# ---------- Column names ----------
col_names <- c(
  "Basin","Loc1","Loc2","Loc1.2",
  rep(c("Mean","SD","N"), 9)  # 9 metric blocks (3 groups × 3 metrics)
)

kableExtra::kbl(
  display_tbl,
  booktabs = TRUE,
  align = align_vec,
  col.names = col_names,
  caption = "Parrot community & functional-group summary at loc2 (Mean, SD, N)."
) %>%
  kableExtra::add_header_above(
    c(" " = 4,
      "Density" = 3, "Biomass" = 3, "Bioerosion" = 3,       # Community
      "Density" = 3, "Biomass" = 3, "Bioerosion" = 3,       # Scrapers
      "Density" = 3, "Biomass" = 3, "Bioerosion" = 3),      # Excavators
    bold = FALSE
  ) %>%
  kableExtra::add_header_above(
    c(" " = 4, "Community" = 9, "Scrapers" = 9, "Excavators" = 9)
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 11,
    bootstrap_options = c("striped","hover","condensed","responsive")
  ) %>%
  kableExtra::scroll_box(width = "100%", height = "520px")
```



## Environmental data

### Summary statistics for loc2 and loc3 sites

```{r environmental_stats_loc2_loc3, echo=FALSE, message=FALSE, warning=FALSE}

# Clean environmental variables once
environmentalData_clean <- environmentalData %>%
  mutate(
    mean_Sea_Surface_Temperature_CRW_daily_01mo = as.numeric(mean_Sea_Surface_Temperature_CRW_daily_01mo),
    mean_Sea_Surface_Salinity_MirasSMOS_3Day_03mo = as.numeric(mean_Sea_Surface_Salinity_MirasSMOS_3Day_03mo),
    mean_PAR_NASA_VIIRS_monthly_01mo = as.numeric(mean_PAR_NASA_VIIRS_monthly_01mo),
    mean_Chlorophyll_A_ESA_OC_CCI_v6.0_monthly_01mo = as.numeric(mean_Chlorophyll_A_ESA_OC_CCI_v6.0_monthly_01mo)
  ) %>%
  filter(!is.na(mean_Sea_Surface_Salinity_MirasSMOS_3Day_03mo)) %>%
  filter(!is.na(mean_PAR_NASA_VIIRS_monthly_01mo)) %>%
  filter(!is.na(mean_Chlorophyll_A_ESA_OC_CCI_v6.0_monthly_01mo))

# Summary by loc2
environmentalStats_loc2 <- environmentalData_clean %>%
  group_by(basin, loc1, loc2) %>%
  summarise(
    maxDate = max(date, na.rm = TRUE),
    minDate = min(date, na.rm = TRUE),
    n = n(),

    maxTemperature = max(mean_Sea_Surface_Temperature_CRW_daily_01mo, na.rm = TRUE),
    minTemperature = min(mean_Sea_Surface_Temperature_CRW_daily_01mo, na.rm = TRUE),
    meanTemperature = mean(mean_Sea_Surface_Temperature_CRW_daily_01mo, na.rm = TRUE),

    maxSalinity = max(mean_Sea_Surface_Salinity_MirasSMOS_3Day_03mo, na.rm = TRUE),
    minSalinity = min(mean_Sea_Surface_Salinity_MirasSMOS_3Day_03mo, na.rm = TRUE),
    meanSalinity = mean(mean_Sea_Surface_Salinity_MirasSMOS_3Day_03mo, na.rm = TRUE),

    maxPAR = max(mean_PAR_NASA_VIIRS_monthly_01mo, na.rm = TRUE),
    minPAR = min(mean_PAR_NASA_VIIRS_monthly_01mo, na.rm = TRUE),
    meanPAR = mean(mean_PAR_NASA_VIIRS_monthly_01mo, na.rm = TRUE),

    maxChlorophyll = max(mean_Chlorophyll_A_ESA_OC_CCI_v6.0_monthly_01mo, na.rm = TRUE),
    minChlorophyll = min(mean_Chlorophyll_A_ESA_OC_CCI_v6.0_monthly_01mo, na.rm = TRUE),
    meanChlorophyll = mean(mean_Chlorophyll_A_ESA_OC_CCI_v6.0_monthly_01mo, na.rm = TRUE),

    .groups = "drop"
  )

# Summary by loc3
environmentalStats_loc3 <- environmentalData_clean %>%
  group_by(basin, loc1, loc2, loc1.2, loc3, lat, lon) %>%
  summarise(
    maxDate = max(date, na.rm = TRUE),
    minDate = min(date, na.rm = TRUE),
    n = n(),

    maxTemperature = max(mean_Sea_Surface_Temperature_CRW_daily_01mo, na.rm = TRUE),
    minTemperature = min(mean_Sea_Surface_Temperature_CRW_daily_01mo, na.rm = TRUE),
    meanTemperature = mean(mean_Sea_Surface_Temperature_CRW_daily_01mo, na.rm = TRUE),

    maxSalinity = max(mean_Sea_Surface_Salinity_MirasSMOS_3Day_03mo, na.rm = TRUE),
    minSalinity = min(mean_Sea_Surface_Salinity_MirasSMOS_3Day_03mo, na.rm = TRUE),
    meanSalinity = mean(mean_Sea_Surface_Salinity_MirasSMOS_3Day_03mo, na.rm = TRUE),

    maxPAR = max(mean_PAR_NASA_VIIRS_monthly_01mo, na.rm = TRUE),
    minPAR = min(mean_PAR_NASA_VIIRS_monthly_01mo, na.rm = TRUE),
    meanPAR = mean(mean_PAR_NASA_VIIRS_monthly_01mo, na.rm = TRUE),

    maxChlorophyll = max(mean_Chlorophyll_A_ESA_OC_CCI_v6.0_monthly_01mo, na.rm = TRUE),
    minChlorophyll = min(mean_Chlorophyll_A_ESA_OC_CCI_v6.0_monthly_01mo, na.rm = TRUE),
    meanChlorophyll = mean(mean_Chlorophyll_A_ESA_OC_CCI_v6.0_monthly_01mo, na.rm = TRUE),

    .groups = "drop"
  )

# Display loc2 table
kbl(
  environmentalStats_loc2,
  caption = "Environmental summary statistics by loc2"
) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  scroll_box(height = "500px")

# Display loc3 table
kbl(
  environmentalStats_loc3,
  caption = "Environmental summary statistics by loc3"
) %>%
  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
  scroll_box(height = "500px")
```

### AragSatState by loc2
```{r aragSatStateLoc2Summary, echo=FALSE}

# Load bottle data
bottle_dir <- file.path(rootFolder, "data", "bottleData")
bottleLoc2_path <- file.path(bottle_dir, "bottleData_BMUs_loc2.csv")

bottleData_loc2 <- readr::read_csv(
  bottleLoc2_path,
  show_col_types = FALSE
)

# Summarize aragonite saturation state by loc2
aragSatState_loc2 <- bottleData_loc2 %>%
  dplyr::group_by(basin, loc1, loc2) %>%
  dplyr::summarise(
    n_bottle = sum(!is.na(AragSatState)),
    mean_AragSatState = mean(AragSatState, na.rm = TRUE),
    sd_AragSatState = sd(AragSatState, na.rm = TRUE),
    se_AragSatState = sd_AragSatState / sqrt(n_bottle),
    min_AragSatState = min(AragSatState, na.rm = TRUE),
    max_AragSatState = max(AragSatState, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    dplyr::across(
      where(is.numeric),
      ~ dplyr::if_else(is.nan(.x) | is.infinite(.x), NA_real_, .x)
    )
  )

# Display table
aragSatState_loc2
```






## Urban influence data
## Calculates distance to nearest city directly at the loc3 level.
Note Loc2 is not done due to it being the average of multiple loc3 and therefore artificially not on a specific site
```{r humanPressureProximityData_loc3, fig.width=12, fig.height=18, out.width="100%", echo=FALSE, message=FALSE, warning=FALSE}

# Natural Earth city/populated-place layer
cities <- ne_download(
  scale = 10,
  type = "populated_places",
  category = "cultural",
  returnclass = "sf"
)

# Helper function to add nearest city name and distance
add_nearest_city <- function(dat) {
  
  dat_sf <- dat %>%
    filter(!is.na(latitude), !is.na(longitude)) %>%
    st_as_sf(
      coords = c("longitude", "latitude"),
      crs = 4326,
      remove = FALSE
    )
  
  nearest_city_idx <- st_nearest_feature(dat_sf, cities)
  
  dist_city_km <- as.numeric(
    st_distance(
      dat_sf,
      cities[nearest_city_idx, ],
      by_element = TRUE
    )
  ) / 1000
  
  dat_sf %>%
    st_drop_geometry() %>%
    mutate(
      closest_city = cities$NAME[nearest_city_idx],
      dist_city_km = dist_city_km
    )
}

# Build one row per loc3 using mean loc3 coordinates
loc3_points <- barData %>%
  select(basin, loc1, loc2, loc3, latitude, longitude) %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  group_by(basin, loc1, loc2, loc3) %>%
  summarise(
    latitude = mean(latitude, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE),
    n_bar_rows = n(),
    n_unique_gps = n_distinct(paste(latitude, longitude)),
    .groups = "drop"
  )

# Add nearest city directly to loc3 points
humanStats_loc3 <- add_nearest_city(loc3_points) %>%
  select(
    basin, loc1, loc2, loc3,
    latitude, longitude,
    closest_city, dist_city_km,
    n_bar_rows, n_unique_gps
  ) %>%
  arrange(basin, loc1, loc2, loc3)

# Display loc3 closest city table
humanStats_loc3 %>%
  kbl(
    digits = 2,
    caption = "Closest city to each loc3 site"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

## Combine datasets
Combine bar stats and parrot community stats tables
```{r, echo=FALSE}
 combinedTableLoc2 <- barStatsLoc2 %>%
  inner_join(
    parrot_loc2_summary,
    by = c("basin", "loc1", "loc2")
  ) %>%
  inner_join(
    aragSatState_loc2,
    by = c("basin", "loc1", "loc2")
  ) %>%
  inner_join(
    environmentalStats_loc2,
    by = c("basin", "loc1", "loc2")
  )

 combinedTableLoc3 <- barStatsLoc3 %>% #note that this table does not have the aragonite saturation state data because not appropriate to group at that level
  inner_join(
    parrot_loc3_summary,
    by = c("basin", "loc1", "loc2", "loc3")
  ) %>%
  inner_join(
    environmentalStats_loc3,
    by = c("basin", "loc1", "loc2", "loc3")
  ) %>%
  inner_join(
    humanStats_loc3,
    by = c("basin", "loc1", "loc2", "loc3")
  )
```




# Vertical dotplots for map by loc1 and loc2
```{r dotplots_by_loc1_loc2_means, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# ===== Metrics =====
panel1_vars <- c(
  "accretion_mm3cm2y",
  "macroboring_mm3cm2y",
  "grazing_mm3cm2y"
)

metric_labels <- c(
  accretion_mm3cm2y   = "Accretion",
  macroboring_mm3cm2y = "Macroboring",
  grazing_mm3cm2y     = "Grazing"
)

volume_var <- "deltaVolume_mm3cm2y"

metric_labels_full <- c(
  metric_labels,
  setNames("Total volume", volume_var)
)

# ===== GLOBAL SHARED X LIMITS =====
shared_xlim <- range(
  c(
    barData$accretion_mm3cm2y,
    barData$macroboring_mm3cm2y,
    barData$grazing_mm3cm2y,
    barData[[volume_var]]
  ),
  na.rm = TRUE
)

# ===== Metric colors =====
metric_colors <- c(
  "Accretion"    = "#2E8B57",
  "Macroboring"  = "#E69F00",
  "Grazing"      = "#C44E52",
  "Total volume" = "#222222"
)

# Vertical offsets within each loc2 row
offset_map <- c(
  "Grazing"      = -0.15,
  "Macroboring"  = -0.05,
  "Accretion"    =  0.05,
  "Total volume" =  0.15
)

# ===== Output folders =====
out_dir <- "regionPlots_loc1_loc2"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

img_dir <- file.path(out_dir, "report_imgs")
if (!dir.exists(img_dir)) dir.create(img_dir, recursive = TRUE)

# Base width and height scaling
base_width <- 7              # inches
base_height_for_three <- 7   # 3 loc2 -> height 7 in

# ===== Loop over loc1 regions =====
for (this_loc1 in sort(unique(barData$loc1))) {

  df_loc1 <- barData %>%
    dplyr::filter(loc1 == this_loc1)

  if (nrow(df_loc1) == 0) next

  # ---------- Long format, then summarise to loc2-level mean + SD ----------
  df_long <- df_loc1 %>%
    dplyr::select(loc2, dplyr::all_of(c(panel1_vars, volume_var))) %>%
    tidyr::pivot_longer(
      cols      = -loc2,
      names_to  = "metric",
      values_to = "value"
    ) %>%
    dplyr::group_by(loc2, metric) %>%
    dplyr::summarise(
      mean = mean(value, na.rm = TRUE),
      sd   = sd(value,   na.rm = TRUE),
      n    = sum(!is.na(value)),
      .groups = "drop"
    ) %>%
    dplyr::filter(!is.na(mean))

  if (nrow(df_long) == 0) next

  # ---------- Metric labels & factor ordering ----------
  df_long <- df_long %>%
    dplyr::mutate(
      metric_label = metric_labels_full[metric],
      metric_label = factor(
        metric_label,
        levels = c("Accretion", "Macroboring", "Grazing", "Total volume")
      )
    )

  # ---------- Base y positions for loc2 with custom ordering per loc1 ----------
  loc2_levels <- sort(unique(df_long$loc2))

  # Caribbean: Puerto Rico above SaintCroix
  if (grepl("carib", this_loc1, ignore.case = TRUE)) {
    idx_pr <- which(grepl("puerto", loc2_levels, ignore.case = TRUE))
    idx_sc <- which(grepl("croix",  loc2_levels, ignore.case = TRUE))
    if (length(idx_pr) == 1 && length(idx_sc) == 1 && idx_pr < idx_sc) {
      tmp <- loc2_levels[idx_pr]
      loc2_levels[idx_pr] <- loc2_levels[idx_sc]
      loc2_levels[idx_sc] <- tmp
    }
  }

  # Marianas: bottom → top = Guam, Saipan, Pagan, Maug
  if (grepl("maria", this_loc1, ignore.case = TRUE)) {
    desired <- c("Guam", "Saipan", "Pagan", "Maug")
    match_one <- function(target) {
      hit <- grepl(target, loc2_levels, ignore.case = TRUE)
      loc2_levels[hit]
    }
    ordered_hits <- unlist(lapply(desired, match_one))
    ordered_hits <- unique(ordered_hits)
    others <- setdiff(loc2_levels, ordered_hits)
    loc2_levels <- c(ordered_hits, others)
  }

  # Samoa: bottom → top = Tutuila, Tau, Ofu
  if (grepl("samoa", this_loc1, ignore.case = TRUE)) {
    desired <- c("Tutuila", "Tau", "Ofu")
    match_one <- function(target) {
      hit <- grepl(target, loc2_levels, ignore.case = TRUE)
      loc2_levels[hit]
    }
    ordered_hits <- unlist(lapply(desired, match_one))
    ordered_hits <- unique(ordered_hits)
    others <- setdiff(loc2_levels, ordered_hits)
    loc2_levels <- c(ordered_hits, others)
  }

  df_long <- df_long %>%
    dplyr::mutate(
      loc2_factor = factor(loc2, levels = loc2_levels),
      y_base      = as.numeric(loc2_factor),
      y_pos       = y_base + offset_map[as.character(metric_label)],
      sd          = ifelse(is.na(sd), 0, sd)
    )

  # ---------- Pretty loc2 labels ----------
  loc2_labels_pretty <- stringr::str_to_title(loc2_levels)

  # ---------- Pretty loc1 title using explicit mapping ----------
  loc1_pretty <- dplyr::recode(
    this_loc1,
    "caribbean"                = "Caribbean",
    "florida"                  = "Florida",
    "flowerGardenBanks"        = "Flower Garden Banks",
    "marianas"                 = "Marianas",
    "middleHawaiianIslands"    = "Middle Hawaiian Islands",
    "northwestHawaiianIslands" = "Northwest Hawaiian Islands",
    "pacificRemoteIslandArea"  = "Pacific Remote Island Area",
    "samoa"                    = "American Samoa",
    .default                   = this_loc1
  )

  # ---------- Plot ----------
  p_loc1 <- ggplot2::ggplot(
    df_long,
    ggplot2::aes(x = mean, y = y_pos, color = metric_label)
  ) +
    ggplot2::geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.7) +
    ggplot2::geom_errorbarh(
      data = dplyr::filter(df_long, metric_label != "Total volume"),
      ggplot2::aes(xmin = mean - sd, xmax = mean + sd, y = y_pos),
      height = 0, linewidth = 0.8
    ) +
    ggplot2::geom_errorbarh(
      data = dplyr::filter(df_long, metric_label == "Total volume"),
      ggplot2::aes(xmin = mean - sd, xmax = mean + sd, y = y_pos),
      height = 0.08, linewidth = 1.6
    ) +
    ggplot2::geom_point(
      ggplot2::aes(size = ifelse(metric_label == "Total volume", 5.5, 3.8)),
      alpha = 0.9
    ) +
    ggplot2::scale_color_manual(name = NULL, values = metric_colors) +
    ggplot2::scale_size_identity() +
    ggplot2::coord_cartesian(xlim = shared_xlim) +
    ggplot2::scale_y_continuous(
      breaks = seq_along(loc2_levels),
      labels = loc2_labels_pretty
    ) +
    ggplot2::xlab(expression(Delta~volume~(mm^3~cm^{-2}~y^{-1}))) +
    ggplot2::ylab(NULL) +
    ggplot2::ggtitle(loc1_pretty) +
    ggplot2::theme_minimal(base_size = 28) +
    ggplot2::theme(
      axis.text.x      = ggplot2::element_text(size = 24),
      axis.text.y      = ggplot2::element_text(size = 22),
      axis.title.x     = ggplot2::element_text(size = 28),
      axis.title.y     = ggplot2::element_blank(),
      plot.title       = ggplot2::element_text(size = 30, face = "bold", hjust = 0.5),
      panel.grid.major = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank(),
      panel.border     = ggplot2::element_rect(color = "black", fill = NA, linewidth = 0.8),
      panel.background = ggplot2::element_rect(fill = "white", color = NA),
      legend.position  = "none",
      plot.margin      = ggplot2::margin(12, 12, 12, 12)
    )

  # ---------- Dynamic height based on number of loc2 ----------
  n_loc2  <- length(loc2_levels)
  out_hgt <- base_height_for_three * (n_loc2 / 3)

  # Optional caps to prevent ridiculous tall/short figures
  out_hgt <- max(out_hgt, 5)
  out_hgt <- min(out_hgt, 16)

  # ---------- Save PNG for report and include it (THIS is what makes heights vary) ----------
  img_file <- file.path(img_dir, paste0("regionPlot_loc1_", this_loc1, ".png"))

  ggplot2::ggsave(
    filename = img_file,
    plot     = p_loc1,
    width    = base_width,
    height   = out_hgt,
    units    = "in",
    dpi      = 200
  )

  # Include in knitted doc (width fixed; height varies via aspect ratio)
  cat("\n\n")
  cat(knitr::asis_output(paste0("![](", img_file, "){width=100%}\n\n")))
  cat("\n\n")

  # ---------- Save to PDF (your original artifact) ----------
  out_pdf <- file.path(out_dir, paste0("regionPlot_loc1_", this_loc1, ".pdf"))
  ggplot2::ggsave(
    filename = out_pdf,
    plot     = p_loc1,
    width    = base_width,
    height   = out_hgt,
    units    = "in",
    device   = cairo_pdf
  )
}
```






#Drivers of carbonate persistence:
Data were analyzed using linear models. Separate models were constructed for each basin, at two spatial scales, island-level and site-level, depending on the spatial resolution of the predictor data. Response parameters (Δmass, Δvolume, Δdensity, accretion, grazing, and macroboring) were grouped at the appropriate scale, resulting in an island or site-specific mean, used in the subsequent analysis. Candidate predictors were first analyzed for collinearity and models were constructed to maintain adequate residual degrees of freedom, and avoid strongly collinear predictors. 

The effects of parrotfish grazing pressure and urban influence were analyzed at the site scale, with depth included as covariate in the Pacific. In the Atlantic, only parrotfish grazing pressure was used as a predictor due to strong collinearity among predictors, partly reflecting the lower number of Atlantic sites relative to the Pacific (8 vs. 37; Table SX; Figure SXX). Environmental predictors were evaluated at the island level, with chlorophyll and aragonite saturation state considered in the Atlantic, and depth added to the two parameters in the Pacific. Temperature (satellite-derived) was not evaluated due to collinearity with aragonite saturation state and mean site depth in both basins.  

## Check collinearity among parrotfish pressure, Urban Influence, and depth
```{r loc3_predictor_collinearity_table, echo=FALSE, message=FALSE, warning=FALSE}

# Build loc3 predictor dataset
loc3_predictor_dat <- combinedTableLoc3 %>%
  select(
    basin, loc1, loc2, loc3,
    comm_bioerosion_mean,
    dist_city_km,
    meanDepth
  ) %>%
  rename(
    `Parrotfish pressure` = comm_bioerosion_mean,
    `Urban Influence` = dist_city_km,
    Depth = meanDepth
  ) %>%
  mutate(
    Basin = recode(basin, atlantic = "Atlantic", pacific = "Pacific")
  ) %>%
  filter(
    !is.na(`Parrotfish pressure`),
    !is.na(`Urban Influence`),
    !is.na(Depth)
  )

# Define pairwise predictor comparisons
pairwise_collinearity_dat <- bind_rows(
  loc3_predictor_dat %>%
    transmute(
      Basin,
      comparison = "Parrotfish pressure vs. Urban Influence",
      x = `Urban Influence`,
      y = `Parrotfish pressure`
    ),
  
  loc3_predictor_dat %>%
    transmute(
      Basin,
      comparison = "Parrotfish pressure vs. Depth",
      x = Depth,
      y = `Parrotfish pressure`
    ),
  
  loc3_predictor_dat %>%
    transmute(
      Basin,
      comparison = "Urban Influence vs. Depth",
      x = Depth,
      y = `Urban Influence`
    )
)

# Correlation table by Basin and predictor pair
loc3_collinearity_cor <- pairwise_collinearity_dat %>%
  group_by(Basin, comparison) %>%
  summarise(
    n_loc3 = n(),
    correlation = cor(x, y, use = "complete.obs"),
    p.value = cor.test(x, y)$p.value,
    .groups = "drop"
  )

# Display correlation table
loc3_collinearity_cor %>%
  mutate(
    correlation = round(correlation, 3),
    p.value = signif(p.value, 3)
  ) %>%
  kbl(
    caption = "Collinearity among loc3-level predictors by Basin",
    col.names = c(
      "Basin",
      "Predictor comparison",
      "n",
      "Correlation",
      "p-value"
    )
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```



## Analyze response variables at the loc3 level 
 Loc3-level analysis using existing combinedTableLoc3
 Each loc3 row is one statistical replicate.

 Separate models are run for Atlantic and Pacific.
 Atlantic: response ~ parrotfish pressure only
 Pacific:  response ~ parrotfish pressure + urban influence + depth
```{r combinedTableLoc3_basin_specific_models, echo=FALSE, message=FALSE, warning=FALSE}
# Define loc3-level BAR response variables in the desired output order
response_vars <- c(
  "meanDeltaMass_mgcm2y",
  "meanDeltaVolume",
  "meanDeltaDensity",
  "meanAccretion",
  "meanGrazing",
  "meanMacroboring"
)

# Response variable display labels with HTML superscripts
response_labels <- c(
  meanDeltaMass_mgcm2y = "Mass (mg cm<sup>-2</sup> yr<sup>-1</sup>)",
  meanDeltaVolume      = "Volume (mm<sup>3</sup> cm<sup>-2</sup> yr<sup>-1</sup>)",
  meanDeltaDensity     = "Density (mg cm<sup>-3</sup>cm<sup>-2</sup> yr<sup>-1</sup>)",
  meanAccretion        = "Accretion (mm<sup>3</sup> cm<sup>-2</sup> yr<sup>-1</sup>)",
  meanGrazing          = "Grazing (mm<sup>3</sup> cm<sup>-2</sup> yr<sup>-1</sup>)",
  meanMacroboring      = "Macroboring (mm<sup>3</sup> cm<sup>-2</sup> yr<sup>-1</sup>)"
)

# Define model term order for the full coefficient table
term_order <- c(
  "(Intercept)",
  "comm_bioerosion_mean",
  "dist_city_km",
  "meanDepth"
)

# Model term display labels
term_labels <- c(
  `(Intercept)`        = "Intercept",
  comm_bioerosion_mean = "Parrotfish pressure",
  dist_city_km         = "Urban influence",
  meanDepth            = "Depth"
)

# Build model data
loc3_model_long <- combinedTableLoc3 %>%
  select(
    basin, loc1, loc2, loc3,
    comm_bioerosion_mean,
    dist_city_km,
    meanDepth,
    all_of(response_vars)
  ) %>%
  pivot_longer(
    cols = all_of(response_vars),
    names_to = "response",
    values_to = "mean_response"
  ) %>%
  mutate(
    response = factor(response, levels = response_vars)
  ) %>%
  filter(
    basin %in% c("atlantic", "pacific"),
    !is.na(mean_response),
    !is.na(comm_bioerosion_mean)
  )

# Function to apply basin-specific NA filtering
filter_model_data <- function(dat, loc3ModelBasin) {
  if (loc3ModelBasin == "atlantic") {
    dat %>%
      filter(!is.na(comm_bioerosion_mean))
  } else if (loc3ModelBasin == "pacific") {
    dat %>%
      filter(
        !is.na(comm_bioerosion_mean),
        !is.na(dist_city_km),
        !is.na(meanDepth)
      )
  } else {
    stop("Unexpected basin name: ", loc3ModelBasin)
  }
}

# Function to fit basin-specific loc3-level models
# Atlantic: response ~ parrotfish pressure only
# Pacific:  response ~ parrotfish pressure + urban influence + depth
fit_loc3_lm <- function(dat, loc3ModelBasin) {
  if (loc3ModelBasin == "atlantic") {
    lm(
      mean_response ~ comm_bioerosion_mean,
      data = dat
    )
  } else if (loc3ModelBasin == "pacific") {
    lm(
      mean_response ~ comm_bioerosion_mean + dist_city_km + meanDepth,
      data = dat
    )
  } else {
    stop("Unexpected basin name: ", loc3ModelBasin)
  }
}

# Function to extract model coefficients and model fit statistics
extract_lm_effects <- function(dat, loc3ModelBasin) {
  model_summary <- summary(fit_loc3_lm(dat, loc3ModelBasin))

  as.data.frame(model_summary$coefficients) %>%
    rownames_to_column("term") %>%
    as_tibble() %>%
    mutate(
      r_squared = model_summary$r.squared,
      adj_r_squared = model_summary$adj.r.squared
    )
}

# Fit models and create full coefficient table
full_model_effects <- loc3_model_long %>%
  group_by(basin, response) %>%
  nest() %>%
  mutate(
    model_data = map2(data, basin, filter_model_data),
    n_loc3 = map_int(model_data, nrow),
    n_loc2 = map_int(model_data, ~ n_distinct(.x$loc2)),
    n_loc1 = map_int(model_data, ~ n_distinct(.x$loc1)),
    coef_table = map2(model_data, basin, extract_lm_effects)
  ) %>%
  select(
    basin, response,
    n_loc3, n_loc2, n_loc1,
    coef_table
  ) %>%
  unnest(coef_table) %>%
  rename(
    estimate  = Estimate,
    std.error = `Std. Error`,
    statistic = `t value`,
    p.value   = `Pr(>|t|)`
  ) %>%
  mutate(
    response = factor(response, levels = response_vars),
    term = factor(term, levels = term_order)
  ) %>%
  select(
    basin, response,
    n_loc3, n_loc2, n_loc1,
    r_squared, adj_r_squared,
    term, estimate, std.error, statistic, p.value
  ) %>%
  arrange(response, basin, term) %>%
  ungroup()

# Create display table with formatted labels and capitalization
full_model_effects_display <- full_model_effects %>%
  mutate(
    Response = response_labels[as.character(response)],
    Basin = str_to_title(basin),
    Term = term_labels[as.character(term)]
  ) %>%
  select(
    Response, Basin,
    n_loc3, n_loc2, n_loc1,
    r_squared, adj_r_squared,
    Term, estimate, std.error, statistic, p.value
  ) %>%
  rename(
    `N loc3` = n_loc3,
    `N loc2` = n_loc2,
    `N loc1` = n_loc1,
    `R-squared` = r_squared,
    `Adjusted R-squared` = adj_r_squared,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t value` = statistic,
    `p value` = p.value
  )

# Identify significant non-intercept term rows
sig_rows <- which(
  full_model_effects$p.value < 0.05 &
    as.character(full_model_effects$term) != "(Intercept)"
)

# Table of full model effects
full_model_effects_display %>%
  kbl(
    digits = 4,
    escape = FALSE,
    caption = "Full loc3-level model coefficient table using combinedTableLoc3. Atlantic models include parrotfish pressure only. Pacific models include parrotfish pressure, urban influence, and depth."
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  row_spec(
    sig_rows,
    bold = TRUE
  ) %>%
  collapse_rows(
    columns = 1,
    valign = "top",
    latex_hline = "major"
  ) %>%
  scroll_box(height = "1500px", width = "100%")
```


```{r loc3_parrotfish_pressure_model_prediction_plots, echo=FALSE, message=FALSE, warning=FALSE}

library(tidyverse)
library(broom)
library(patchwork)

# Responses to plot, in the requested order
response_lookup <- tibble(
  response_var = c("meanGrazing", "meanAccretion", "meanDeltaVolume"),
  response_label = c("Grazing", "Accretion", "Volume")
)

y_axis_labels <- list(
  meanGrazing = expression("Grazing (mm"^3*" cm"^-2*" yr"^-1*")"),
  meanAccretion = expression("Accretion (mm"^3*" cm"^-2*" yr"^-1*")"),
  meanDeltaVolume = expression("Volume (mm"^3*" cm"^-2*" yr"^-1*")")
)

# Consistent basin colors
basin_colors <- c(
  Atlantic = "#D55E00",
  Pacific = "#0072B2"
)

basin_shapes <- c(
  Atlantic = 16,
  Pacific = 17
)

# Build long-format plotting/modeling dataset
loc3_parrotfish_dat <- combinedTableLoc3 %>%
  select(
    basin, loc1, loc2, loc3,
    comm_bioerosion_mean,
    dist_city_km,
    meanDepth,
    all_of(response_lookup$response_var)
  ) %>%
  pivot_longer(
    cols = all_of(response_lookup$response_var),
    names_to = "response_var",
    values_to = "value"
  ) %>%
  left_join(response_lookup, by = "response_var") %>%
  mutate(
    basin = str_to_title(basin)
  ) %>%
  filter(
    basin %in% c("Atlantic", "Pacific"),
    !is.na(value),
    !is.na(comm_bioerosion_mean)
  )

# Fit basin-specific models
# Atlantic: response ~ parrotfish pressure
# Pacific: response ~ parrotfish pressure + urban influence + depth
model_fits <- loc3_parrotfish_dat %>%
  group_by(response_var, response_label, basin) %>%
  nest() %>%
  mutate(
    fit = map2(
      data,
      basin,
      ~ if (.y == "Atlantic") {
        lm(value ~ comm_bioerosion_mean, data = .x)
      } else {
        lm(value ~ comm_bioerosion_mean + dist_city_km + meanDepth, data = .x)
      }
    )
  )

# Extract parrotfish-pressure p values
parrotfish_pvals <- model_fits %>%
  mutate(tidy_fit = map(fit, tidy)) %>%
  select(response_var, response_label, basin, tidy_fit) %>%
  unnest(tidy_fit) %>%
  filter(term == "comm_bioerosion_mean") %>%
  mutate(
    significant = p.value < 0.05,
    p_label = case_when(
      p.value < 0.001 ~ paste0(basin, ": p < 0.001"),
      TRUE ~ paste0(basin, ": p = ", formatC(p.value, format = "f", digits = 3))
    )
  )

# Build prediction data
# For Pacific, urban influence and depth are held constant at Pacific means
prediction_dat <- model_fits %>%
  mutate(
    pred_dat = map2(
      data,
      basin,
      ~ {
        x_seq <- seq(
          min(.x$comm_bioerosion_mean, na.rm = TRUE),
          max(.x$comm_bioerosion_mean, na.rm = TRUE),
          length.out = 200
        )
        
        if (.y == "Atlantic") {
          tibble(
            comm_bioerosion_mean = x_seq
          )
        } else {
          tibble(
            comm_bioerosion_mean = x_seq,
            dist_city_km = mean(.x$dist_city_km, na.rm = TRUE),
            meanDepth = mean(.x$meanDepth, na.rm = TRUE)
          )
        }
      }
    ),
    predictions = map2(
      fit,
      pred_dat,
      ~ bind_cols(
        .y,
        as_tibble(
          predict(
            .x,
            newdata = .y,
            interval = "confidence",
            level = 0.95
          )
        )
      )
    )
  ) %>%
  select(response_var, response_label, basin, predictions) %>%
  unnest(predictions) %>%
  left_join(
    parrotfish_pvals %>%
      select(response_var, basin, significant),
    by = c("response_var", "basin")
  ) %>%
  filter(significant)

# Fixed p-value label positions in lower right so they do not overlap
label_dat <- parrotfish_pvals %>%
  left_join(
    loc3_parrotfish_dat %>%
      group_by(response_var, response_label) %>%
      summarize(
        x_min = min(comm_bioerosion_mean, na.rm = TRUE),
        x_max = max(comm_bioerosion_mean, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("response_var", "response_label")
  ) %>%
  mutate(
    x = x_max - 0.02 * (x_max - x_min),
    basin = factor(basin, levels = c("Atlantic", "Pacific"))
  ) %>%
  arrange(response_var, basin) %>%
  group_by(response_var, response_label) %>%
  mutate(
    y = case_when(
      row_number() == 1 ~ -75,
      row_number() == 2 ~ -90,
      TRUE ~ -90
    )
  ) %>%
  ungroup()

# Plotting function
plot_response <- function(response_var_name) {
  
  response_title <- response_lookup %>%
    filter(response_var == response_var_name) %>%
    pull(response_label)
  
  y_lab <- y_axis_labels[[response_var_name]]
  
  point_dat <- loc3_parrotfish_dat %>%
    filter(response_var == response_var_name)
  
  line_dat <- prediction_dat %>%
    filter(response_var == response_var_name)
  
  label_subdat <- label_dat %>%
    filter(response_var == response_var_name)
  
  ggplot() +
    geom_hline(
      yintercept = 0,
      linetype = "dashed",
      color = "black",
      linewidth = 0.6
    ) +
    geom_point(
      data = point_dat,
      aes(
        x = comm_bioerosion_mean,
        y = value,
        color = basin,
        shape = basin
      ),
      size = 2.6,
      alpha = 0.80
    ) +
    geom_ribbon(
      data = line_dat,
      aes(
        x = comm_bioerosion_mean,
        ymin = lwr,
        ymax = upr,
        fill = basin,
        group = basin
      ),
      alpha = 0.18,
      color = NA
    ) +
    geom_line(
      data = line_dat,
      aes(
        x = comm_bioerosion_mean,
        y = fit,
        color = basin,
        group = basin
      ),
      linewidth = 1
    ) +
    geom_label(
      data = label_subdat,
      aes(
        x = x,
        y = y,
        label = p_label,
        color = basin
      ),
      hjust = 1,
      vjust = 0,
      size = 3,
      label.size = NA,
      fill = "white",
      alpha = 0.85,
      show.legend = FALSE
    ) +
    scale_color_manual(values = basin_colors) +
    scale_fill_manual(values = basin_colors) +
    scale_shape_manual(values = basin_shapes) +
    coord_cartesian(ylim = c(-100, 50)) +
    scale_y_continuous(breaks = seq(-100, 50, by = 25)) +
    labs(
      x = "Parrotfish pressure",
      y = y_lab,
      title = response_title,
      color = "Basin",
      fill = "Basin",
      shape = "Basin"
    ) +
    theme_classic() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      panel.border = element_rect(color = "black", fill = NA),
      legend.position = "bottom"
    )
}

# Build plots in requested order
grazing_plot <- plot_response("meanGrazing")
accretion_plot <- plot_response("meanAccretion")
volume_plot <- plot_response("meanDeltaVolume")

# Arrange panels
grazing_plot + accretion_plot + volume_plot +
  plot_layout(ncol = 3, guides = "collect") &
  theme(legend.position = "bottom")
```




```{r loc3_urban_influence_model_prediction_plots, echo=FALSE, message=FALSE, warning=FALSE}

library(tidyverse)
library(broom)
library(patchwork)

# Responses to plot
response_lookup <- tibble(
  response_var = c("meanGrazing", "meanAccretion", "meanDeltaVolume", "meanMacroboring"),
  response_label = c("Grazing", "Accretion", "Volume", "Macroboring")
)

y_axis_labels <- list(
  meanGrazing = expression("Grazing (mm"^3*" cm"^-2*" yr"^-1*")"),
  meanAccretion = expression("Accretion (mm"^3*" cm"^-2*" yr"^-1*")"),
  meanDeltaVolume = expression("Volume (mm"^3*" cm"^-2*" yr"^-1*")"),
  meanMacroboring = expression("Macroboring (mm"^3*" cm"^-2*" yr"^-1*")")
)

# Pacific-only dataset
loc3_urban_dat <- combinedTableLoc3 %>%
  select(
    basin, loc1, loc2, loc3,
    comm_bioerosion_mean,
    dist_city_km,
    meanDepth,
    all_of(response_lookup$response_var)
  ) %>%
  filter(basin == "pacific") %>%
  pivot_longer(
    cols = all_of(response_lookup$response_var),
    names_to = "response_var",
    values_to = "value"
  ) %>%
  left_join(response_lookup, by = "response_var") %>%
  filter(
    !is.na(value),
    !is.na(dist_city_km),
    !is.na(comm_bioerosion_mean),
    !is.na(meanDepth)
  )

# Fit Pacific models
# response ~ parrotfish pressure + urban influence + depth
model_fits <- loc3_urban_dat %>%
  group_by(response_var, response_label) %>%
  nest() %>%
  mutate(
    fit = map(
      data,
      ~ lm(value ~ comm_bioerosion_mean + dist_city_km + meanDepth, data = .x)
    )
  )

# Extract urban-influence p values
urban_pvals <- model_fits %>%
  mutate(tidy_fit = map(fit, tidy)) %>%
  select(response_var, response_label, tidy_fit) %>%
  unnest(tidy_fit) %>%
  filter(term == "dist_city_km") %>%
  mutate(
    significant = p.value < 0.05,
    p_label = case_when(
      p.value < 0.001 ~ "p < 0.001",
      TRUE ~ paste0("p = ", formatC(p.value, format = "f", digits = 3))
    )
  )

# Build prediction data
# Hold parrotfish pressure and depth constant at Pacific means
prediction_dat <- model_fits %>%
  mutate(
    pred_dat = map(
      data,
      ~ tibble(
        dist_city_km = seq(
          min(.x$dist_city_km, na.rm = TRUE),
          max(.x$dist_city_km, na.rm = TRUE),
          length.out = 200
        ),
        comm_bioerosion_mean = mean(.x$comm_bioerosion_mean, na.rm = TRUE),
        meanDepth = mean(.x$meanDepth, na.rm = TRUE)
      )
    ),
    predictions = map2(
      fit,
      pred_dat,
      ~ bind_cols(
        .y,
        as_tibble(
          predict(
            .x,
            newdata = .y,
            interval = "confidence",
            level = 0.95
          )
        )
      )
    )
  ) %>%
  select(response_var, response_label, predictions) %>%
  unnest(predictions)

# Lower-right p-value positions for each plot
label_dat <- loc3_urban_dat %>%
  group_by(response_var, response_label) %>%
  summarize(
    x_min = min(dist_city_km, na.rm = TRUE),
    x_max = max(dist_city_km, na.rm = TRUE),
    y_min = min(value, na.rm = TRUE),
    y_max = max(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(urban_pvals, by = c("response_var", "response_label")) %>%
  mutate(
    x = x_max - 0.02 * (x_max - x_min),
    y = y_min + 0.06 * (y_max - y_min)
  )

# Plotting function
plot_response <- function(response_var_name) {
  
  response_title <- response_lookup %>%
    filter(response_var == response_var_name) %>%
    pull(response_label)
  
  y_lab <- y_axis_labels[[response_var_name]]
  
  point_dat <- loc3_urban_dat %>%
    filter(response_var == response_var_name)
  
  line_dat <- prediction_dat %>%
    filter(response_var == response_var_name)
  
  label_subdat <- label_dat %>%
    filter(response_var == response_var_name)
  
  ggplot() +
    geom_hline(
      yintercept = 0,
      linetype = "dashed",
      color = "black",
      linewidth = 0.6
    ) +
    geom_point(
      data = point_dat,
      aes(
        x = dist_city_km,
        y = value
      ),
      size = 2.6,
      alpha = 0.80
    ) +
    geom_ribbon(
      data = line_dat,
      aes(
        x = dist_city_km,
        ymin = lwr,
        ymax = upr
      ),
      alpha = 0.18
    ) +
    geom_line(
      data = line_dat,
      aes(
        x = dist_city_km,
        y = fit
      ),
      linewidth = 1
    ) +
    geom_label(
      data = label_subdat,
      aes(
        x = x,
        y = y,
        label = p_label
      ),
      hjust = 1,
      vjust = 0,
      size = 3,
      label.size = NA,
      fill = "white",
      alpha = 0.85
    ) +
    labs(
      x = "Distance to nearest city (km)",
      y = y_lab,
      title = response_title
    ) +
    theme_classic() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      panel.border = element_rect(color = "black", fill = NA)
    )
}

# Build plots in preferred order
grazing_plot <- plot_response("meanGrazing")
accretion_plot <- plot_response("meanAccretion")
volume_plot <- plot_response("meanDeltaVolume")
macroboring_plot <- plot_response("meanMacroboring")

# Arrange panels
(grazing_plot + accretion_plot) /
(volume_plot + macroboring_plot)
```


















# Run linear models for Loc2 site-level analysis 
Linear models were used in each basin to investigate the impact of aragonite saturation state on BAR responses. Models were run on loc2-averaged values, with depth and chlorophyll included as covariates. After depth and chlorophyll were accounted for, omega relationships were evaluated separately for each response variable and basin.

Note sites like St. Thomas and Wake with no density are not included in the analysis only of those parameters, as listed with NAs.

This is what ChatGPT told me about removing depth (don't use):
"Depth was not a predictor of interest, it did not improve interpretation, and including it costs degrees of freedom in a small loc2-level dataset."


## Are loc2 environmental predictors collinear?
```{r loc2_environmental_predictor_collinearity_table, echo=FALSE, message=FALSE, warning=FALSE}
# Build loc2 predictor dataset
loc2_predictor_dat <- combinedTableLoc2 %>%
  select(
    basin, loc1, loc2,
    meanChlorophyll,
    meanTemperature,
    meanDepth,
    mean_AragSatState
  ) %>%
  rename(
    Chlorophyll = meanChlorophyll,
    Temperature = meanTemperature,
    Depth = meanDepth,
    Omega = mean_AragSatState
  ) %>%
  filter(
    basin %in% c("atlantic", "pacific"),
    !is.na(Chlorophyll),
    !is.na(Temperature),
    !is.na(Depth),
    !is.na(Omega)
  )

# Define pairwise predictor comparisons
pairwise_collinearity_dat <- bind_rows(
  loc2_predictor_dat %>%
    transmute(
      basin, loc1, loc2,
      comparison = "Omega vs. Chlorophyll",
      x = Chlorophyll,
      y = Omega
    ),

  loc2_predictor_dat %>%
    transmute(
      basin, loc1, loc2,
      comparison = "Omega vs. Temperature",
      x = Temperature,
      y = Omega
    ),

  loc2_predictor_dat %>%
    transmute(
      basin, loc1, loc2,
      comparison = "Omega vs. Depth",
      x = Depth,
      y = Omega
    ),

  loc2_predictor_dat %>%
    transmute(
      basin, loc1, loc2,
      comparison = "Chlorophyll vs. Temperature",
      x = Temperature,
      y = Chlorophyll
    ),

  loc2_predictor_dat %>%
    transmute(
      basin, loc1, loc2,
      comparison = "Chlorophyll vs. Depth",
      x = Depth,
      y = Chlorophyll
    ),

  loc2_predictor_dat %>%
    transmute(
      basin, loc1, loc2,
      comparison = "Temperature vs. Depth",
      x = Depth,
      y = Temperature
    )
)

# Correlation table by basin and predictor pair
loc2_collinearity_cor <- pairwise_collinearity_dat %>%
  group_by(basin, comparison) %>%
  summarise(
    n = n(),
    correlation = cor(x, y, use = "complete.obs"),
    p.value = cor.test(x, y)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    Basin = str_to_title(basin),
    Comparison = comparison
  ) %>%
  select(
    Basin,
    Comparison,
    n,
    correlation,
    p.value
  ) %>%
  rename(
    Correlation = correlation,
    `p value` = p.value
  ) %>%
  arrange(Comparison, Basin)

# Display correlation table
loc2_collinearity_cor %>%
  kbl(
    digits = 3,
    caption = "Pairwise correlations among loc2-level environmental predictors."
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  collapse_rows(
    columns = 1,
    valign = "top",
    latex_hline = "major"
  )
```

## Analyze response variables at the loc2 level
```{r lmLoc2ByBasinChlorophyllOmegaDepth_cleanTables, echo=FALSE, message=FALSE, warning=FALSE}
# Define loc2-level BAR response variables in the desired output order
response_vars <- c(
  "meanDeltaMass_mgcm2y",
  "meanDeltaVolume",
  "meanDeltaDensity",
  "meanAccretion",
  "meanGrazing",
  "meanMacroboring"
)

# Response variable display labels with HTML superscripts
response_labels <- c(
  meanDeltaMass_mgcm2y = "Mass (mg cm<sup>-2</sup> yr<sup>-1</sup>)",
  meanDeltaVolume      = "Volume (mm<sup>3</sup> cm<sup>-2</sup> yr<sup>-1</sup>)",
  meanDeltaDensity     = "Density (mg cm<sup>-3</sup>)",
  meanAccretion        = "Accretion (mm<sup>3</sup> cm<sup>-2</sup> yr<sup>-1</sup>)",
  meanGrazing          = "Grazing (mm<sup>3</sup> cm<sup>-2</sup> yr<sup>-1</sup>)",
  meanMacroboring      = "Macroboring (mm<sup>3</sup> cm<sup>-2</sup> yr<sup>-1</sup>)"
)

# Define model term order for the full coefficient table
term_order_loc2 <- c(
  "(Intercept)",
  "mean_AragSatState",
  "meanChlorophyll",
  "meanDepth"
)

# Model term display labels
term_labels_loc2 <- c(
  `(Intercept)`       = "Intercept",
  mean_AragSatState   = "Omega",
  meanChlorophyll     = "Chlorophyll",
  meanDepth           = "Depth"
)

# Put the data in long format where "response" is the name of the response variable
# and "value" is the value of that response variable.
model_dat_long <- combinedTableLoc2 %>%
  select(
    basin, loc1, loc2,
    mean_AragSatState,
    meanChlorophyll,
    meanDepth,
    all_of(response_vars)
  ) %>%
  pivot_longer(
    cols = all_of(response_vars),
    names_to = "response",
    values_to = "value"
  ) %>%
  mutate(
    response = factor(response, levels = response_vars)
  ) %>%
  filter(
    basin %in% c("atlantic", "pacific"),
    !is.na(value),
    !is.na(mean_AragSatState),
    !is.na(meanChlorophyll)
  )

# Function to apply basin-specific NA filtering
filter_model_data <- function(dat, loc2ModelBasin) {
  if (loc2ModelBasin == "atlantic") {
    dat %>%
      filter(
        !is.na(mean_AragSatState),
        !is.na(meanChlorophyll)
      )
  } else if (loc2ModelBasin == "pacific") {
    dat %>%
      filter(
        !is.na(mean_AragSatState),
        !is.na(meanChlorophyll),
        !is.na(meanDepth)
      )
  } else {
    stop("Unexpected basin name: ", loc2ModelBasin)
  }
}

# Function to fit basin-specific loc2-level models
# Atlantic: response ~ omega + chlorophyll
# Pacific:  response ~ omega + chlorophyll + depth
fit_loc2_lm <- function(dat, loc2ModelBasin) {
  if (loc2ModelBasin == "atlantic") {
    lm(
      value ~ mean_AragSatState + meanChlorophyll,
      data = dat
    )
  } else if (loc2ModelBasin == "pacific") {
    lm(
      value ~ mean_AragSatState + meanChlorophyll + meanDepth,
      data = dat
    )
  } else {
    stop("Unexpected basin name: ", loc2ModelBasin)
  }
}

# Function to extract model coefficients and model fit statistics
extract_lm_effects <- function(dat, loc2ModelBasin) {
  model_summary <- summary(fit_loc2_lm(dat, loc2ModelBasin))

  as.data.frame(model_summary$coefficients) %>%
    rownames_to_column("term") %>%
    as_tibble() %>%
    mutate(
      r_squared = model_summary$r.squared,
      adj_r_squared = model_summary$adj.r.squared
    )
}

# Fit models and create full coefficient table
full_model_effects_loc2 <- model_dat_long %>%
  group_by(basin, response) %>%
  nest() %>%
  mutate(
    model_data = map2(data, basin, filter_model_data),
    n_loc2 = map_int(model_data, nrow),
    n_loc1 = map_int(model_data, ~ n_distinct(.x$loc1)),
    coef_table = map2(model_data, basin, extract_lm_effects)
  ) %>%
  select(
    basin, response,
    n_loc2, n_loc1,
    coef_table
  ) %>%
  unnest(coef_table) %>%
  rename(
    estimate  = Estimate,
    std.error = `Std. Error`,
    statistic = `t value`,
    p.value   = `Pr(>|t|)`
  ) %>%
  mutate(
    response = factor(response, levels = response_vars),
    term = factor(term, levels = term_order_loc2)
  ) %>%
  select(
    basin, response,
    n_loc2, n_loc1,
    r_squared, adj_r_squared,
    term, estimate, std.error, statistic, p.value
  ) %>%
  arrange(response, basin, term) %>%
  ungroup()

# Create display table with formatted labels and capitalization
full_model_effects_loc2_display <- full_model_effects_loc2 %>%
  mutate(
    Response = response_labels[as.character(response)],
    Basin = str_to_title(basin),
    Term = term_labels_loc2[as.character(term)]
  ) %>%
  select(
    Response, Basin,
    n_loc2, n_loc1,
    r_squared, adj_r_squared,
    Term, estimate, std.error, statistic, p.value
  ) %>%
  rename(
    `N loc2` = n_loc2,
    `N loc1` = n_loc1,
    `R-squared` = r_squared,
    `Adjusted R-squared` = adj_r_squared,
    Estimate = estimate,
    `Std. Error` = std.error,
    `t value` = statistic,
    `p value` = p.value
  )

# Identify significant non-intercept term rows
sig_rows <- which(
  full_model_effects_loc2$`p.value` < 0.05 &
    as.character(full_model_effects_loc2$term) != "(Intercept)"
)

# Full model effects table
full_model_effects_loc2_display %>%
  kbl(
    digits = 4,
    escape = FALSE,
    caption = "Full loc2-level model coefficient table using combinedTableLoc2. Atlantic models include omega and chlorophyll. Pacific models include omega, chlorophyll, and depth."
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  row_spec(
    sig_rows,
    bold = TRUE
  ) %>%
  collapse_rows(
    columns = 1,
    valign = "top",
    latex_hline = "major"
  ) %>%
  scroll_box(height = "1500px", width = "100%")
```

